---
title: "CHRIS untargeted metabolomics: alignment and correspondence analysis"
author: "Johannes Rainer"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
---

```{r biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown()
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dev = c("png", "pdf"))
```

**Modified**: `r file.info("alignment_correspondence.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE}
## Set general options
options(useFancyQuotes = FALSE)
set.seed(123)

## Define paths:
FILE_NAME <- "alignment_correspondence"
## Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", FILE_NAME, "/")
if (dir.exists(IMAGE_PATH)) unlink(IMAGE_PATH, recursive = TRUE, force = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE, showWarnings = FALSE)
## Path to store RData files
RDATA_PATH <- paste0("data/RData/", FILE_NAME, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

## Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 6))

rt_cut <- 340

MZML_PATH <- "/data/massspec/mzML/"
if (!dir.exists(MZML_PATH))
    stop("Can not find the directory with the mzML files: ", MZML_PATH)

```

- Test first all on *just* the QC samples.
- perform alignment
  - subset alignment based on QC samples: QC samples get aligned against each
    other and then study samples get aligned against QC samples.
- perform correspondence
- plot EICs for standards in:
  - x random samples
  - 2 QC samples per batch
- peak filling

# Introduction

This document describes the alignment and correspondence analysis of the
HILIC-based untargeted metabolomics data from the CHRIS population study.

In a first pilot analysis alignment and correspondence is performed *only* on
the QC samples of the study.

The chromatographic peak detection was defined in and performed by the
[peak_detection.Rmd](peak_detection.Rmd) file.

Below we load the data and all required libraries.

```{r libs}
library(xcms)

register(bpstart(MulticoreParam(ncores - 1L)))

load("data/RData/peak_detection/data_pos.RData")
dr <- dirname(data_pos)
dirname(data_pos) <- sub("/Volumes/extdata/data/mzML/",
                         "/data/massspec/mzML/", dr)

data_qc <- filterFile(data_pos, which(data_pos$type == "QC"))
```


# Lab-internal standards

We next load the lab-internal standards. EICs on these will be used to evaluate
the raw data as well as the alignment and correspondence results.

```{r stds}
## Extract known compunds
library("MetaboCoreUtils")
library(Rdisop)
std_info <- read.table(
    "https://raw.githubusercontent.com/EuracBiomedicalResearch/lcms-standards/master/data/standards_dilution.txt",
    sep = "\t", header = TRUE, as.is = TRUE)
std_info <- std_info[!is.na(std_info[, "POS"]), ]
rownames(std_info) <- 1:nrow(std_info)
std_info$mzneut = NA
std_info$mz_ion = NA
for (i in seq(nrow(std_info))) {
    if (grepl("C", std_info$formula[i])) {
        std_info$mzneut[i] <- getMolecule(
            as.character(std_info$formula[i]))$exactmass
    } else {
        std_info$mzneut[i] = as.numeric(std_info$formula[i])
    }
    ## Calculate also the m/z
    std_info$mz_ion[i] <- mass2mz(
        std_info$mzneut[i], adduct = std_info[i, "POS"])[1, 1]
}
std_info <- std_info[!is.na(std_info$mz_ion), ]
std_info <- std_info[order(std_info$name), ]
```


# Alignment

Due to the large shifts in the retention times, alignment (using the default
settings for our LC-MS setup) did not work properly. Thus we try to align the
batches first using a manually selected set of standards and their retention
times in QC samples across batches.

```{r}
std_selected <- c(
    "1-Methylhistidine",                # and 3-Methylhistidine
    "8-Oxo-2-Deoxyguanosine",
    "Acetylhistidine",
    "Adenine",
    "ADMA",
    "Alanine",
    "Arginine",
    "Asparagine",
    ## "Betaine",                          # no clear, single peak.
    "C5 Carnitine",
    "Caffeine",
    "cGMP",
    "Citrulline",
    "Creatine",
    "Cystine",
    "Dihydroxyacetone phosphate",
    "Fructose",
    "Glutamine",
    "Glyceraldehyde 2-phosphate",
    "Glycero-phosphocholine",
    "Glycine",
    "Histidine",
    "Hydroxyproline",
    "Hypoxanthine",                     # take the second peak
    "Indoleacetic acid",
    "Inosine",
    "L-Glutamic Acid",
    "Lysine",
    "Methionine",
    "N-Acetylornithine",
    "Ornithine",
    "Proline",
    "Putrescine",
    "SDMA",
    "Serine",
    "Sphingosine",
    "Taurine",
    "Threonine"
)

std_rt <- std_info[std_info$name %in% std_selected, ]

```

We next identify chromatographic peaks for each of these standards and plot
EICs.

```{r match-standards}
library(MetaboAnnotation)
param <- MzRtParam(tolerance = 0, ppm = 20, toleranceRt = 50) # rt 40 before
std_peaks <- matchMz(std_rt, chromPeaks(data_qc), param = param,
                     mzColname = c("mz_ion", "mz"),
                     rtColname = c("RT", "rt"))
std_peaks <- std_peaks[whichQuery(std_peaks)]
std_rt <- query(std_peaks)

## For each standard, define m/z and rt ranges.
std_rt_mz <- do.call(rbind, lapply(seq_along(std_peaks), function(z) {
    tmp <- std_peaks[z]
    c(range(tmp$target_rtmin, tmp$target_rtmax),
      range(tmp$target_mzmin, tmp$target_mzmax))
}))
colnames(std_rt_mz) <- c("rtmin", "rtmax", "mzmin", "mzmax")

system.time(
    std_chr <- chromatogram(data_qc, rt = std_rt_mz[, c("rtmin", "rtmax")],
                            mz = std_rt_mz[, c("mzmin", "mzmax")])
)

```

```{r raw-eic-plots, echo = FALSE}
#' Plot EICs only for a single batch.
plot_single_sample <- function(x, col = "#00000060", rt = 0) {
    par(mfrow = c(ncol(x), 1), mar = c(2, 4, 1, 0.5))
    smpl <- basename(x$mzML_file)
    for (i in seq_len(ncol(x))) {
        tmp <- x[, i]
        plot(tmp, main = smpl[i], col = col, peakCol = col,
             peakBg = paste0(col, 20))
        abline(v = rt, lty = 2)
        abline(v = chromPeaks(tmp)[, "rt"], col = col)
        grid()
        legend("topright", legend = i, cex = 2)
    }
}

plot_split_batch <- function(x, col, name = "", rt = 0) {
    btches <- unique(x$batch)
    par(mfrow = c(length(btches), 1), mar = c(0, 4, 0, 0.5))
    for (btch in btches) {
        plot(x[, x$batch == btch], col = col[btch], xlab = "", main = "",
             yaxt = "n", ylab = "", col.axis = "#00000060",
             peakType = "none")
        abline(v = rt, lty = 2)
        grid()
        legend("topright", btch)
    }
}

col_batch <- rainbow(length(unique(data_qc$batch)))
names(col_batch) <- unique(data_qc$batch)
col_sample <- col_batch[data_qc$batch]
dr <- paste0(IMAGE_PATH, "standards-alignment-raw/")
dir.create(dr, showWarnings = FALSE)

rt_assign <- data.frame()

for (i in seq_len(nrow(std_rt))) {
    ## Overview plot
    pdf(paste0(dr, std_rt$name[i], "-all-batches.pdf"), width = 8, height = 5)
    chr <- std_chr[i, ]
    plot(chr, col = paste0(col_sample, 80), peakType = "point",
         peakCol = paste0(col_sample[chromPeaks(chr)[, "sample"]], 40))
    grid()
    abline(v = std_rt$RT[i], lty = 2)
    legend("topright", c(std_rt$name[i], paste0("rt = ", std_rt$RT[i])))
    dev.off()
    ## One plot, split by batch
    pdf(paste0(dr, std_rt$name[i], "-per-batch.pdf"), width = 8, height = 30)
    plot_split_batch(chr, col = col_batch, name = std_rt$name[i],
                     rt = std_rt$RT[i])
    dev.off()
    ## Individual samples, one plot per batch
    dir.create(paste0(dr, std_rt$name[i]), showWarnings = FALSE)
    for (batch in names(col_batch)) {
        pdf(paste0(dr, std_rt$name[i], "/", batch, ".pdf"), width = 8,
            height = 10)
        chr_batch <- chr[, chr$batch == batch]
        plot_single_sample(chr_batch, col = col_batch[batch],
                           rt = std_rt$RT[i])
        dev.off()
        ## define data.frame to export
        fls <- basename(pData(chr_batch)$mzML_file)
        rt_batch <- data.frame(name = std_rt$name[i], batch = batch,
                               index = seq_along(fls), rt = 0,
                               mzML_file = fls)
        pks <- chromPeaks(chr_batch)
        if (nrow(pks)) {
            add_tmp <- data.frame(name = std_rt$name[i], batch = batch,
                                  index = pks[, "column"], rt = pks[, "rt"],
                                  mzML_file = fls[pks[, "column"]])
            rt_batch <- rbind(rt_batch[!rt_batch$index %in% add_tmp$index, ],
                            add_tmp)
            rt_batch <- rt_batch[order(rt_batch$index), ]
        }
        rt_assign <- rbind(rt_assign, rt_batch)
    }
}

library(writexl)
write_xlsx(as.data.frame(rt_assign), path = "data/_temp_alignment_rt.xlsx")

## Continue with cGMP
## Notes:
## - Alanine, batches 2017-20 and 2017-22: shifts in the same 2 positions.
## - Alanine: batches 2020_10, 2020_11, 2020_14, 2020_15 long tails
## - Arginine: 2020_21 10 sec shift left.
## - Asparagine: 2020_18 -> 10 sec shift left.
## - C5 Carnitine: 2020_19 -> looking at wrong ion? large shift left;
##   2017-91, 92, many overlapping peaks
```

- Notes for peak detection: lower snthresh or minimum required signal. merge
  peaks: maybe increase rt range?

Alignment is performed on the pooled QC samples using the default settings for
the employed LC-MS setup.

```{r alignment, echo = TRUE, message = FALSE}
## Grouping the peaks according to group
pdp_subs <- PeakDensityParam(
    sampleGroups = data_pos$type, bw = 3,
    minFraction = 0.8, binSize = 0.02, maxFeatures = 200)
data_pos <- groupChromPeaks(data_pos, param = pdp_subs)
## Subset alignment options
pgp_subs <- PeakGroupsParam(minFraction = 2/3,
                            subset = which(data_pos$type == "QC"),
                            subsetAdjust = "previous", span = 0.5,
                            extraPeaks = 100)
## Perform the alignment
system.time(
    data_pos <- adjustRtime(data_pos, param = pgp_subs)
)
save(data_pos, file = paste0(RDATA_PATH, "data_pos_align.RData"))
```

We next plot the retention times for the selected *hook peaks*.

```{r alignment-hook-peak-rt, fig.path = IMAGE_PATH, fig.cap = "Retention times of hook peaks across the various QC samples.", echo = FALSE}
load(paste0(RDATA_PATH, "data_pos_align.RData"))

data_qc <- filterFile(data_pos, which(data_pos$type == "QC"))
## Define a color for each batch.
col_batch <- rainbow(length(unique(data_qc$batch)))
names(col_batch) <- unique(data_qc$batch)
## Plot the hook peaks
pgm <- peakGroupsMatrix(processParam(data_qc@.processHistory[[3]]))

plot(3, 3, pch = NA, xlim = range(pgm, na.rm = TRUE), ylim = c(1, nrow(pgm)),
     xlab = "retention time", ylab = "peak number")
for (i in seq_len(nrow(pgm))) {
    points(x = pgm[i, ], y = rep(i, ncol(pgm)),
           col = paste0(col_batch[data_qc$batch], 10), pch = 16)
}
grid()

```

Hook peaks are mostly present at around 30 seconds as well as after 150 (up to
220) seconds. Retention time shifts become larger with higher retention
times. The mean retention time range is `r format(mean(apply(pgm, MARGIN = 1, function(z) diff(range(z, na.rm = TRUE)))), digits = 3)`.

We next plot the retention time differences for QC samples.

```{r alignment-adjustrtime-plot, fig.path = IMAGE_PATH, fig.cap = "Alignment results", echo = FALSE}
tmp <- data_qc
tmp@.processHistory <- list()
plotAdjustedRtime(tmp, col = paste0(col_batch[tmp$batch], 20))
```

Matching identified chromatographic peaks with standards to create EIC plots.

```{r known-cmps, message = FALSE, warning = FALSE}

library(MetaboAnnotation)
param <- MzRtParam(tolerance = 0, ppm = 20, toleranceRt = 40)
std_peaks <- matchMz(std_info, chromPeaks(data_qc), param = param,
                     mzColname = c("mz_ion", "mz"),
                     rtColname = c("RT", "rt"))
std_peaks <- std_peaks[whichQuery(std_peaks)]
std_info <- query(std_peaks)

## For each standard, define m/z and rt ranges.
std_rt_mz <- do.call(rbind, lapply(seq_along(std_peaks), function(z) {
    tmp <- std_peaks[z]
    c(range(tmp$target_rtmin, tmp$target_rtmax),
      range(tmp$target_mzmin, tmp$target_mzmax))
}))
colnames(std_rt_mz) <- c("rtmin", "rtmax", "mzmin", "mzmax")

system.time(
std_chr <- chromatogram(data_qc, rt = std_rt_mz[, c("rtmin", "rtmax")],
                        mz = std_rt_mz[, c("mzmin", "mzmax")],
                        adjustedRtime = TRUE)
)
system.time(
    std_chr_raw <- chromatogram(dropAdjustedRtime(data_qc),
                                rt = std_rt_mz[, c("rtmin", "rtmax")],
                                mz = std_rt_mz[, c("mzmin", "mzmax")])
)

## Plot these friends




dr <- paste0(IMAGE_PATH, "standards-alignment-aligned/")
dir.create(dr, showWarnings = FALSE)
history <- processHistory(data_pos)
save(history, file = paste0(dr, "history.RData"))
for (i in seq_len(nrow(std_info))) {
    pdf(paste0(dr, std_info$name[i], "-all-batches.pdf"), width = 8, height = 5)
    chr <- std_chr[i, ]
    plot(chr, col = paste0(col_sample, 80), peakType = "point",
         peakCol = paste0(col_sample[chromPeaks(chr)[, "sample"]], 40))
    grid()
    abline(v = std_info$RT[i], lty = 2)
    legend("topright", c(std_info$name[i], paste0("rt = ", std_info$RT[i])))
    dev.off()
    pdf(paste0(dr, std_info$name[i], "-per-batch.pdf"), width = 8, height = 30)
    plot_batch(chr, col = col_batch, name = std_info$name[i],
               rt = std_info$RT[i])
    dev.off()
}

dr <- paste0(IMAGE_PATH, "standards-alignment-raw/")
dir.create(dr, showWarnings = FALSE)
for (i in seq_len(nrow(std_info))) {
    pdf(paste0(dr, std_info$name[i], "-all-batches.pdf"), width = 8, height = 5)
    chr <- std_chr_raw[i, ]
    plot(chr, col = paste0(col_sample, 80), peakType = "point",
         peakCol = paste0(col_sample[chromPeaks(chr)[, "sample"]], 40))
    grid()
    abline(v = std_info$RT[i], lty = 2)
    legend("topright", c(std_info$name[i], paste0("rt = ", std_info$RT[i])))
    dev.off()
    pdf(paste0(dr, std_info$name[i], "-per-batch.pdf"), width = 8, height = 30)
    plot_batch(chr, col = col_batch, name = std_info$name[i],
               rt = std_info$RT[i])
    dev.off()
}

## Manually selected standards with ~ OK (and unambiguous) data.
std_selected <- c(
    "1-Methylhistidine",                # and 3-Methylhistidine
    "8-Oxo-2-Deoxyguanosine",
    "Acetylhistidine",
    "Adenine",
    "ADMA",
    "Alanine",
    "Arginine",
    "Asparagine",
    "Betaine",
    "C5 Carnitine",
    "Caffeine",
    "cGMP",
    "Citrulline",
    "Creatine",
    "Cystine",
    "Dihydroxyacetone phosphate",
    "Fructose",
    "Glutamine",
    "Glyceraldehyde 2-phosphate",
    "Glycero-phosphocholine",
    "Glycine",
    "Histidine",
    "Hydroxyproline",
    "Hypoxanthine",                     # take the second peak
    "Indoleacetic acid",
    "Inosine",
    "L-Glutamic Acid",
    "Lysine",
    "Methionine",
    "N-Acetylornithine",
    "Ornithine",
    "Proline",
    "Putrescine",
    "SDMA",
    "Serine",
    "Sphingosine",
    "Taurine",
    "Threonine"
)


```

TODO:
- select standards with the potential to yield ~ OK signal.
- for selected standards, record their rt in QC samples of each batch.
- Need to plot the data separately for each batch, each QC sample to record it.
- Ideally, determine the retention time for matching peaks per sample, export as
  xlsx and manually adjust/fix them.



# Correspondence

We next perform the correspondence analysis on the aligned data.

```{r}
pdp <- PeakDensityParam(
    sampleGroups = data_pos$type, bw = 2,
    minFraction = 0.3, binSize = 0.02, maxFeatures = 200)
data_pos <- groupChromPeaks(data_pos, param = pdp)
```

# Gap filling


# Session information

```{r}
sessionInfo()
```

# References
