---
title: "CHRIS untargeted metabolomics: alignment and correspondence analysis"
author: "Johannes Rainer"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
---

```{r biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown()
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

**Modified**: `r file.info("alignment_correspondence.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE}
## Set general options
options(useFancyQuotes = FALSE)
set.seed(123)

## Define paths:
FILE_NAME <- "alignment_correspondence"
## Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", FILE_NAME, "/")
if (dir.exists(IMAGE_PATH)) unlink(IMAGE_PATH, recursive = TRUE, force = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE, showWarnings = FALSE)
## Path to store RData files
RDATA_PATH <- paste0("data/RData/", FILE_NAME, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

## Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 6))

rt_cut <- 340

MZML_PATH <- "/data/massspec/mzML/"
if (!dir.exists(MZML_PATH))
    stop("Can not find the directory with the mzML files: ", MZML_PATH)

```

- Test first all on *just* the QC samples.
- perform alignment
  - subset alignment based on QC samples: QC samples get aligned against each
    other and then study samples get aligned against QC samples.
- perform correspondence
- plot EICs for standards in:
  - x random samples
  - 2 QC samples per batch
- peak filling

# Introduction

This document describes the alignment and correspondence analysis of the
HILIC-based untargeted metabolomics data from the CHRIS population study.

In a first pilot analysis alignment and correspondence is performed *only* on
the QC samples of the study.

The chromatographic peak detection was defined in and performed by the
[peak_detection.Rmd](peak_detection.Rmd) file.

# Alignment

# Correspondence

# Session information

```{r}
sessionInfo()
```

# References
