---
title: "CHRIS preprocessing posititve mode"
author: "Marilyn De Graeve, Philippine Louail, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-02-28"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
bibliography: references.bib
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("CHRIS_preprocessing_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))

library("BiocStyle")
library("kableExtra")
library("knitr")
suppressMessages(library("rmarkdown"))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
```

# Introduction

During this workflow, the preprocessing (PP) of the untargeted metabolomics data of the Cooperative Health Research in South Tirol (CHRIS) study is performed. For a description of the study, methods used for the collection, handling and aqcuisition of the liquid chromatography-mass spectrometry (LC-MS) samples, please see [@verri_hernandes_age_2022]. Samples are aquired in both positive and negative ionization mode, for which the PP of the positive (pos) mode will be performed in this Rmarkdown document.

# Packages

```{r packages, message=FALSE}
suppressMessages(library(magick))
suppressMessages(library(MetaboAnnotation))
library(MetaboCoreUtils)
library(MsBackendSql)
suppressMessages(library(MsExperiment))
library(MsQuality)
library(pander)
library(pheatmap)
library(RColorBrewer)
library(readxl)
library(RSQLite)
suppressMessages(library(Spectra))
suppressMessages(library(SummarizedExperiment))
suppressMessages(library(vioplot))
suppressMessages(library(xcms))
```



# Session information

```{r}
sessionInfo()
```

# References

