---
title: "CHRIS_remove_dropouts_PP"
author: "Marilyn De Graeve, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-01-15"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

**Modified**: `r file.info("CHRIS_remove_dropouts_PP.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, dev = "png",
               dpi = 300)
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))

```

# Introduction

During this workflow, we aim to identify analytes (semi targeted) co-aqcuired
with the untargeted metabolomics data of the Cooperative Health Research in
South Tirol (CHRIS) study. For a description of the study, methods used for the
collection, handling and aqcuisition of the liquid chromatography-mass
spectrometry (LC-MS) samples, please see [@verri_hernandes_age_2022].  Samples
are acquired in both positive and negative ionization mode, for which the
pre-processing of the positive (pos) mode is performed and discussed in the
Rmarkdown document `CHRIS_preprocessing_pos.Rmd`.  The (feature) filtering and
normalization of the positive (pos) mode is discussed in the Rmarkdown document
`CHRIS_filtering_normalization_pos.Rmd`.  The semi-targeted annotation of the
untargeted metabolomics feature matrix is discussed in the Rmarkdown document
`CHRIS_semi_targeted_annotation`.

From January 2025, from after pre-processing, CHRIS study participant need to be
removed.  The removal of these participant is done in this script. After this,
the `CHRIS_filtering_normalization_pos.Rmd` and
`CHRIS_semi_targeted_annotation.Rmd` scripts will be rerun. After optimization
of the CHRIS PP (without loading everything in memory), the
`CHRIS_preprocessing_pos.Rmd` script will be rerun as well with the updated
sqlite database (01/2025).

More information regarding the **destruction of CHRIS participant data and
samples**:

All participants in the CHRIS study have the right to request destruction of
their data and samples at any time. In addition, they can indicate that their
data and samples should be destroyed after their death. Triggered by an active
request from a CHRIS participant, we have updated the list of CHRIS participants
that dropped out from the study and asked for destruction of their data and
samples.  If you are using CHRIS data and/or samples in your research, you must
immediately destroy all individual level data and samples associated with any of
the above AIDs.  This includes individual level data and samples that you have
received to perform the research activities, but also all individual level data
or samples that you have derived or generated in the process. Previously
generated non-individual level results (for example, regression coefficients
from models run using individual level data) can be kept without modifications.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "CHRIS_remove_dropouts_PP"

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
DATA_PATH <- "data"

```

## Packages

```{r packages, message = FALSE, echo = FALSE}
library(chrisUtils)
library(chrisData)
library(MsExperiment)
library(RSQLite)
library(xcms)
library(Spectra)
library(pander)
library(reshape2)
library(MetaboAnnotation)
library(MetaboCoreUtils)
library(MsBackendSql)
library(SummarizedExperiment)

```


# Load data

Files loaded are the 4 Rdata files created in the `CHRIS_preprocessing_pos.Rmd` script.
Run the code first for the blanks, change the comments and run for the non-blanks
(biological samples and QCs).

XcmsExperiment objects:
!CANNOT be loaded because need connection to the database file! SKIP

```{r load xcms experiment, eval = FALSE}
#' load xcmsexperment
load(file.path(DATA_PATH, "blank_chris.RData")) #blank_sample is xcmsexp object
blank_sample

#load(file.path(DATA_PATH, "samples_chris.RData")) #chris is xcmsexp object, BIG!
#chris

```

SummarizedExperiment objects:

```{r load Rdata}
#' Load preprocessing results
load(file.path(DATA_PATH, "SumExp_chris_blank.RData")) #res_blank is object
res_blank

#load(file.path(DATA_PATH, "SumExp_chris.RData")) #res is object, BIG!
#res

```


# Change name of sqlite db in the SummarizedExperiment object

Jo: the name of the database (version <01/2025) was
"res@spectra@backend@dbname...".  The database name is defined in the
MsBackendOfflineSql backend object, which is in the @backend slot of the Spectra
object, which itself is in the @spectra slot of the MsExperiment (or rather
XcmsExperiment) object res.

res is Summarizedexperiment, object needed (chris, XcmsExperiment) CANNOT be
loaded.

```{r change_db_name}
blank_sample@spectra@backend@dbname  #nok, no acces xcms obj

```


# Remove dropouts

TODO

```{r}

```


# Data export

The data is saved after preprocessing, both as Rdata objects, as well as a
summarized experiment. The data is ready to use for normalization and
(multivariate) statistics.

## Save samples and blank Rdata

Separate blanks from other samples and save as Rdata files.

```{r sep blanks and bio samples}
blank_index <- sampleData(chris)$sample_type == "Blank"
blank_sample <- chris[blank_index, keepAdjustedRtime = TRUE,
                      keepFeatures = TRUE]

#sample_index <- sampleData(chris)$sample_type != "Blank"
#chris <- chris[!blank_index, keepAdjustedRtime = TRUE,
#               keepFeatures = TRUE]

save(blank_sample, file = paste0(RDATA_PATH, "blank_chris.RData"))
#save(chris, file = paste0(RDATA_PATH, "samples_chris.RData"))

```

## Save Summarized experiment

```{r SummarizedExperiment}
#' Extract results as a SummarizedExperiment
res <- quantify(chris, method = "sum", filled = FALSE)
assays(res)$raw_filled <- featureValues(chris, method = "sum", filled = TRUE)
res
save(res, file = paste0(RDATA_PATH, "SumExp_chris.RData"))

# also for blanks (if present)
if (nrow(sampleData(blank_sample)) >= 1) {
    res_blank <- quantify(blank_sample, method = "sum", filled = FALSE)
    assays(res_blank)$raw_filled <- featureValues(blank_sample, method = "sum",
                                                  filled = TRUE )
    save(res_blank, file = paste0(RDATA_PATH, "SumExp_chris_blank.RData"))
    }

```


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```
