---
title: "Mummichog annotation retrieval"
author: "Marilyn De Graeve, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-12-13"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("Mummichog_annotations.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r configuration, include = TRUE, cached = FALSE}

#' introduction
STUDY <- "CHRIS study"
TARGET <- "BMI (numeric value) "
DATASET <- "untargeted metabolomics, positive mode"

#' settings
FILENAME <- "CHRIS_BMI_umetabpos"
LOCATION <- 'local' #'server' OR 'local', change in chunk directories if needed

#' target variables
#print('target of the linear regression model:')
TARGET_COLNAME <- 'BMI'
#print('more concrete defined target of the linear regression model:')
TARGET_LMNAME <- 'BMI'

```


```{r style, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library("BiocStyle")
library("kableExtra") 
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
BiocStyle::markdown()

```

# Introduction

In this document, the Mummichog algorithm from Li et al. are performed on the
significant metabolite list on the targeted metabolomics data after statistical
analyis (linear models or t-test todo!!!) from the CHRIS obesity project.

As POC, we will use the lm results table from:
(metabo_chris_untargeted/CHRIS_lm_pos.html)[metabo_chris_untargeted/CHRIS_lm_pos.html].
There, the exploratory statistical analyis (linear models) are performed on the
untargeted metabolomics data from the CHRIS study (baseline); using the
continious variable "BMI" as a target of interest.

The (static) table with the significant metabolites only contained 11
significant variables, maybe not enough to leverage the power of the Mummichog
algoritm. ??todo with testdata first!!! => NO, input is full list, so also give
the not-significant w their values. 

This document will be TEMPLATE, to be used for different datasets - target
significant tables.


# Setup

## Directories

```{r directories}
#' General settings
filename <- "Mummichog_annotations"

#' Path to the data
DATA_INPUT_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Projects/metabo_chris_untargeted/data' #FINDME!

#' Path to store RData files
DATA_OUTPUT_PATH <- paste0("data/mummichog/")
dir.create(DATA_OUTPUT_PATH, recursive = TRUE, showWarnings = FALSE)


```

## Libraries

```{r libraries}
```

## Functions

```{r functions}
```

# Steps

The following steps are performed:

- [x] in the metabo_chris_untargeted/CHRIS_lm_pos.html: write table with the ALL
  (both significant and non-significant) metabolites with mz, rt, id, p-value (=
  adj.p lm) and statistic (= coef lm) after performing lm with `r TARGET` as
  target, using the `R DATASET` dataset.
- [x] rename colnames so c(m/z, retention_time, p_value, statistic,
  CompoundID_from_user, ...)
- [ ] run mummichog 
  - [x] (now via bash, 
  - [ ] future w spectripy)
- [x] load and cleanup results into 1 df.
- [ ] add results as putative IDs into the sumExp object and save as Rdata.


# load and rename results from lm

load the results from lm analysis. write table with the ALL
(both significant and non-significant) metabolites with mz, rt, id, p-value (=
  adj.p lm) and statistic (= coef lm) after performing lm with `r TARGET` as
  target, using the `R DATASET` dataset.

```{r load-results-lm}

#' example data: 
#    - dataset: umetab pos chris (3820#, come from umExp_chris_filtered_norm.RData) 
#    - target: bmi (float, continious)
#    - covariates: sex, age 
#' after lm: 11 metab signif eg FT01399

#' load data lm-results
handle <- file.path(DATA_PATH, 'lm_output_results_metab.txt')
lm_output_results_metab <- read.csv(handle, header = TRUE, sep="\t")  |>
    as.data.frame()
lm_output_results_metab[1:10, 1:10]
dim(lm_output_results_metab)
```

rename colnames to c(m/z, retention_time, p_value, statistic,
CompoundID_from_user, ...) so it fits the input in mummichog, write the table.

```{r make-write-mummichog-input}
#' make new df with correct names for mummichog
df <- NULL
df$m.z  <- lm_output_results_metab$mzmed
df$retention_time  <- lm_output_results_metab$rtmed
df$p_value  <- lm_output_results_metab[, paste0('p.adj_', TARGET_LMNAME)]
df$statistic  <-lm_output_results_metab[, paste0('coef_', TARGET_LMNAME)]
df$CompoundID_from_user <- lm_output_results_metab$AID
df <- as.data.frame(df)
head(df)

#TODO m/z ipv m.z ...!!!! ok?


#' write the mummichog input table
handle <- file.path(DATA_INPUT_PATH, 'mummichog_input.txt')
write.table(df, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```


# Run mummichog algoritm 

## Run mummichog algoritm using Bash

Before trying to integrate the mummichog algorithm using spectripy, we first
want to run it automatically from this script. it is run locally, on a Linux
computer with the Mummichog installed according to the instructions from
'https://github.com/shuzhao-li/mummichog'.

After install using pip in the linux terminal, it is run using commands:

- cd Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/
- python3 -m mummichog.main -f testdata0710.txt -o t2

the results are saved in the same folder, as the input. you will get a folder
with various output. in the first instance, i.e. for the putative annotation of
untargeted metabolic features, we are interested in 1 tabular output. See
section [Results import] below.

run the command using R (calls bash), for linux:

```{r run-mummichog-from-bash}

#TODO!!!
#command <- 'python3.10 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/testdata0710.txt -o t2' #test
#command <- 'bash ~/Documents/Files/Work_Eurac/Code/Sh_scripts/mummichog2.sh'
command <- 'python3.10 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/data/mummichog_input.txt -o t2'
command <- paste0('python3.10 -m mummichog.main -f ', handle, ' -o t2')
command

system(command)

```

This will be either called in the spectripy environment, if possible. will try
that next. therefor, no need to go too much into win/linux differences (see old
code marilyn if needed).


# Results from mummichog (pasted from html report)

For the CHRIS untargeted metabolomics - pos mode lm model with BMI, 
pasted from the mummichog html report:

User input data

User supplied 3820 features as reference list.We are using 348 features (p < 0.000100) as significant list. The feature level data are shown in the Manhattan plots below.
Top pathways

The reference feature list is mapped to 205 EmpiricalCompounds, significant features to 71 EmpiricalCompounds. In the table below, pathway_size is number of detected EmpiricalCompounds for each pathway; overlap_size is number of significant EmpiricalCompounds. Empirical p-values are estimated by permutation test. Details on EmpiricalCompounds are in 'tables/ListOfEmpiricalCompounds.tsv'.

Pathways 	overlap_size	pathway_size	p-value 	overlap_EmpiricalCompounds
Purine metabolism 	12	17	0.0005	E373,E336,E10,E135,E147,E377,E127,E128,E379,E113,E230,E105
N-Glycan Degradation 	6	6	0.0005	E214,E219,E305,E217,E218,E216
Galactose metabolism 	7	8	0.00067	E214,E219,E305,E339,E217,E218,E216
Sialic acid metabolism 	8	10	0.00084	E214,E219,E305,E339,E217,E51,E218,E216
Alanine and Aspartate Metabolism 	10	14	0.00084	E29,E202,E102,E10,E135,E196,E127,E128,E379,E105
Glycosphingolipid biosynthesis - globoseries 	5	5	0.00092	E214,E219,E217,E218,E216
N-Glycan biosynthesis 	6	7	0.0016	E214,E219,E217,E450,E218,E216
Pyrimidine metabolism 	11	17	0.00176	E29,E102,E84,E135,E49,E111,E127,E85,E128,E379,E105
Caffeine metabolism 	4	4	0.00345	E214,E215,E211,E216
Hexose phosphorylation 	5	6	0.0058	E214,E219,E217,E218,E216
Fructose and mannose metabolism 	5	6	0.0058	E214,E219,E217,E218,E216
Nitrogen metabolism 	5	6	0.0058	E102,E135,E127,E128,E105
Pentose phosphate pathway 	7	10	0.00647	E214,E219,E111,E217,E379,E218,E216
Vitamin B9 (folate) metabolism 	6	8	0.00655	E202,E10,E135,E314,E51,E483
Glycosphingolipid metabolism 	6	8	0.00655	E214,E219,E217,E51,E218,E216
Bile acid biosynthesis 	3	3	0.01151	E379,E10,E5
Vitamin H (biotin) metabolism 	3	3	0.01151	E134,E379,E133
Glycolysis and Gluconeogenesis 	6	9	0.01302	E214,E219,E217,E379,E218,E216
Phosphatidylinositol phosphate metabolism 	6	9	0.01302	E214,E219,E304,E217,E218,E216
Aminosugars metabolism 	4	5	0.01344	E128,E305,E127,E135
Starch and Sucrose Metabolism 	5	7	0.01378	E214,E219,E217,E218,E216
Glycine, serine, alanine and threonine metabolism 	13	27	0.02504	E29,E71,E165,E70,E10,E135,E49,E51,E99,E379,E88,E9,E164
Glutathione Metabolism 	5	8	0.02764	E29,E73,E10,E135,E87
Aspartate and asparagine metabolism 	17	38	0.02966	E73,E379,E105,E133,E102,E240,E134,E10,E135,E49,E91,E127,E92,E128,E87,E62,E196
Vitamin B5 - CoA biosynthesis from pantothenate 	2	2	0.03974	E379,E282
Lipoate metabolism 	2	2	0.03974	E134,E133
CoA Catabolism 	2	2	0.03974	E379,E282
Arachidonic acid metabolism 	3	4	0.04058	E379,E10,E135
Leukotriene metabolism 	3	4	0.04058	E423,E379,E10
Selenoamino acid metabolism 	3	4	0.04058	E29,E379,E51

Top modules

Network modules may capture concerted metabolite activities that are missed by predefined pathways. Use pull-down menus to visualize modules and the activity network, which tries to combine the results of pathway/module analyses. The visualization requires interenet connection.

module_1, p=0.00285, 64 metabolites
Glycine, 2-methylbutyrylglycine, isovalerylglycine, L-Threonine, AMP, Sarcosine; N-Methylglycine, L-Allo-threonine, L-Serine, Hippurate; Hippuric acid; N-Benzoylglycine; Benzoylaminoacetic acid, L-Alanine, N,N-Dimethylglycine; Dimethylglycine, 5-Oxoproline, N-acetyl-L-alanine, 5-L-Glutamyl-L-alanine, L-Glutamate, D-Alanine, D-3-Amino-isobutanoate, beta-Alanine, L-Aspartate, N-Carbamoyl-beta-alanine, 3-Ureidoisobutyrate, 3-Aminoisobutyric acid; 3-Aminoisobutanoate; 3-Amino-2-methylpropanoate, beta-Aminopropion aldehyde, (S)-2-Aminobutanoate; (S)-2-Aminobutanoic acid; (S)-2-Aminobutyric acid, Glycine betaine, 4-Aminobutanoate, L-3-Amino-isobutanoate, D-Serine, D-Proline, L-Proline, L-4-Hydroxyglutamate semialdehyde, L-Valine, 5,6-Dihydrothymine, L-1-Pyrroline-3-hydroxy-5-carboxylate, L-Leucine, L-Isoleucine, L-Tyrosine, L-Asparagine, L-Glutamine, L-Citrulline; 2-Amino-5-ureidovaleric acid; Citrulline, 5-amino-1-(5-phospho-D-ribosyl)imidazole-4-carboxylate, N-Acetyl-L-aspartate, Hypoxanthine, Xanthine, Inosine, 6-Thioxanthine 5'-monophosphate, alpha-N-Phenylacetyl-L-glutamine, L-Lysine, Taurine, 5-Formiminotetrahydrofolate; 5-Formimidoyltetrahydrofolate, N-Formyl-L-glutamate, Tetrahydrofolyl-[Glu](n); Tetrahydrofolyl-[Glu](n+1); THF-polyglutamate; Tetrahydropteroyl-[gamma-Glu]n; Tetrahydropteroyl-[gamma-Glu]n+1, L-2-Aminoadipate; L-alpha-Aminoadipate; L-alpha-Aminoadipic acid; L-2-Aminoadipic acid; L-2-Aminohexanedioate, , D-Glucose, myo-Inositol, D-Mannose, 3-beta-D-Galactosyl-sn-glycerol; Galactosylglycerol, beta-D-Glucose, alpha-D-Glucose, beta-D-Fructose; beta-Fruit sugar; beta-D-arabino-Hexulose; beta-Levulose; Fructose, N-Acetyl-D-glucosamine, N-Acetyl-D-mannosamine; 2-Acetamido-2-deoxy-D-mannose, dGMP
module_2, p=0.00285, 8 metabolites
Acetylhydrazine, 4-Acetamido-2-amino-6-nitrotoluene, Isoniazid, N-Acetylisoniazid; (N)1-Acetylisoniazid, 2,4-Diamino-6-nitrotoluene, N,N'-Diacetylhydrazine; Diacetylhydrazine, 1,7-Dimethylxanthine; Paraxanthine, 5-Acetylamino-6-formylamino-3-methyluracil; AFMU
module_3, p=0.00285, 9 metabolites
AMP, L-Leucine, Taurine, Inosine, L-Lysine, dGMP, Hypoxanthine, Xanthine, L-Valine
module_4, p=0.00285, 6 metabolites
beta-Aminopropion aldehyde, beta-Alanine, 3-Ureidoisobutyrate, 5,6-Dihydrothymine, N-Carbamoyl-beta-alanine, 3-Aminoisobutyric acid; 3-Aminoisobutanoate; 3-Amino-2-methylpropanoate
module_5, p=0.00285, 19 metabolites
L-Tyrosine, (S)-2-Aminobutanoate; (S)-2-Aminobutanoic acid; (S)-2-Aminobutyric acid, L-2-Aminoadipate; L-alpha-Aminoadipate; L-alpha-Aminoadipic acid; L-2-Aminoadipic acid; L-2-Aminohexanedioate, N-Formyl-L-glutamate, 6-Thioxanthine 5'-monophosphate, L-Isoleucine, 5-Formiminotetrahydrofolate; 5-Formimidoyltetrahydrofolate, Tetrahydrofolyl-[Glu](n); Tetrahydrofolyl-[Glu](n+1); THF-polyglutamate; Tetrahydropteroyl-[gamma-Glu]n; Tetrahydropteroyl-[gamma-Glu]n+1, L-Citrulline; 2-Amino-5-ureidovaleric acid; Citrulline, , 4-Aminobutanoate, 5-amino-1-(5-phospho-D-ribosyl)imidazole-4-carboxylate, L-Glutamine, L-3-Amino-isobutanoate, L-Aspartate, N-Acetyl-L-aspartate, L-Glutamate, alpha-N-Phenylacetyl-L-glutamine, L-Asparagine
module_6, p=0.0114, 11 metabolites
L-Allo-threonine, Glycine, L-Threonine, N,N-Dimethylglycine; Dimethylglycine, Hippurate; Hippuric acid; N-Benzoylglycine; Benzoylaminoacetic acid, 2-methylbutyrylglycine, Sarcosine; N-Methylglycine, L-Serine, isovalerylglycine, Glycine betaine, D-Serine
module_7, p=0.0114, 9 metabolites
myo-Inositol, 3-beta-D-Galactosyl-sn-glycerol; Galactosylglycerol, D-Glucose, beta-D-Fructose; beta-Fruit sugar; beta-D-arabino-Hexulose; beta-Levulose; Fructose, N-Acetyl-D-mannosamine; 2-Acetamido-2-deoxy-D-mannose, beta-D-Glucose, N-Acetyl-D-glucosamine, alpha-D-Glucose, D-Mannose
module_8, p=0.01709, 6 metabolites
5-L-Glutamyl-L-alanine, N-acetyl-L-alanine, 5-Oxoproline, D-Alanine, L-Alanine, D-3-Amino-isobutanoate
EmpiricalCompounds favored by above analyses

An EmpiricalCompound is a computational unit for a tentative metabolite. It can group multiple ions, and be any of the isobaric/isomeric species. This table contains EmpiricalCompounds prioritized by pathway/module analysis.

EmpiricalCompound	Input m/z 	Retention time	ion	mz_diff	Statistic 	Significant
E202 	Best guess: C01042, N-Acetyl-L-aspartate	C01045, C01042
	176.055921970663 	173.694713316198	M+H[1+]	0.0005	-0.01	yes
	198.038006551681 	173.861829437395	M+Na[1+]	0.0006	-0.02	yes
E73 	Best guess: C03740, 5-L-Glutamyl-L-alanine	C03740
	219.097532358774 	173.778214366121	M+H[1+]	-0.0	0.01	yes
	241.079235672381 	173.521904078937	M+Na[1+]	-0.0003	0.01	yes
E71 	Best guess: C00188, L-Threonine	C05519, C00188
	120.065523131714 	173.073457066636	M+H[1+]	0.0	-0.0	no
	121.068458062774 	173.073533238983	M(C13)+H[1+]	-0.0004	-0.0	no
	142.047326431032 	173.28763510354	M+Na[1+]	-0.0002	-0.02	yes
E133 	Best guess: C00047, L-Lysine	C00047
	147.112705075048 	188.401597703773	M+H[1+]	-0.0001	0.01	no
	148.115493917227 	188.795383051417	M(C13)+H[1+]	-0.0007	0.01	yes
	169.09429589703 	189.017985302095	M+Na[1+]	-0.0005	0.01	yes
E219 	Best guess: C00159, D-Mannose	C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	203.052605699518 	157.306679354031	M+Na[1+]	-0.0001	0.01	yes
E134 	Best guess: C00047, L-Lysine	C00047
	147.11260202509 	197.510902449481	M+H[1+]	-0.0002	0.01	yes
E10 	Best guess: C00037, Glycine	C00037
	76.0399838804231 	171.000085201666	M+H[1+]	0.0007	-0.02	yes
E135 	Best guess: C05938, L-4-Hydroxyglutamate semialdehyde	C00217, C00025, C05938
	148.060368676706 	173.534733130586	M+H[1+]	-0.0001	0.06	yes
	149.063322293979 	173.717647896661	M(C13)+H[1+]	-0.0006	0.06	yes
	170.041515585337 	173.26346115276	M+Na[1+]	-0.001	0.05	yes
E49 	Best guess: C03284, L-3-Amino-isobutanoate	C02356, C01026, C00334, C05145, C01205, C03284
	104.070626269476 	170.028462115796	M+H[1+]	0.0	0.02	yes
E111 	Best guess: C01801, Deoxyribose	C01801
	135.064793569549 	183.621653668673	M+H[1+]	-0.0004	-0.01	yes
E51 	Best guess: C00740, D-Serine	C00065, C00740
	106.049924871492 	181.214777879227	M+H[1+]	0.0	-0.01	yes
	107.053394356903 	181.220937864838	M(C13)+H[1+]	0.0001	-0.0	no
E5 	Best guess: C00245, Taurine	C00245
	126.021851798413 	171.869322381365	M+H[1+]	-0.0001	-0.0	no
	64.0168155801004 	168.005957346195	M(C13)+2H[2+]	0.0005	-0.01	no
	148.003739920283 	171.759923099613	M+Na[1+]	-0.0002	-0.01	yes
E256 	Best guess: C16618, 6-Thioxanthine 5'-monophosphate	C16618
	381.026323234369 	170.0503149422	M+H[1+]	-0.0002	0.02	yes
	449.011919329513 	169.766617987116	M+HCOONa[1+]	-0.002	0.02	yes
	202.006213859459 	169.561534512327	M+H+Na[2+]	-0.0017	0.01	no
E113 	Best guess: C00262, Hypoxanthine	C01620, C00262
	137.045627869455 	126.582512358947	M+H[1+]	-0.0001	0.01	yes
	138.048236031546 	126.14019835728	M(C13)+H[1+]	-0.0009	0.01	yes
E230 	Best guess: C11821, 5-Hydroxyisourate	C11821
	185.030405750958 	143.799427970004	M+H[1+]	-0.0002	0.03	yes
E164 	Best guess: CE4968, isovalerylglycine	CE4970, CE4968
	181.07179618207 	37.2666769819766	M+Na[1+]	0.0008	0.03	yes
E450 	Best guess: C01246, Dolichyl beta-D-glucosyl phosphate	C01246
	489.225461961012 	183.498298899609	M+Na[1+]	0.003	0.02	yes
E483 	Best guess: C03541, Tetrahydrofolyl-[Glu](n)	C03541
	726.247195793601 	172.170403402053	M+Na[1+]	0.0018	0.02	yes
E423 	Best guess: CE6193, 18-carboxy-dinor-LTE4	CE6193
	440.172943535541 	166.765396065663	M+H[1+]	-0.0008	0.06	yes
E84 	Best guess: C00906, 5,6-Dihydrothymine	C00906
	129.065648761369 	169.888238507885	M+H[1+]	-0.0002	-0.02	yes
E165 	Best guess: CE4968, isovalerylglycine	CE4970, CE4968
	181.071856704847 	49.9074854160341	M+Na[1+]	0.0009	0.04	yes
E147 	Best guess: C00385, Xanthine	C00385
	153.040390069727 	140.221429374424	M+H[1+]	-0.0003	0.02	yes
E85 	Best guess: C00906, 5,6-Dihydrothymine	C00906
	151.047560591964 	179.505557121695	M+Na[1+]	-0.0003	-0.01	yes
E333 	Best guess: C04148, alpha-N-Phenylacetyl-L-glutamine	C04148, C05642
	287.100403998584 	67.1785265203047	M+Na[1+]	0.0001	-0.02	yes
	265.118359707076 	67.2065113808169	M+H[1+]	0.0001	-0.01	yes
E87 	Best guess: C04282, 1-Pyrroline-4-hydroxy-2-carboxylate	C01879, C04281, C04282
	130.049812413726 	179.663907541441	M+H[1+]	-0.0001	-0.01	yes
	130.049763850925 	173.428721779296	M+H[1+]	-0.0001	0.05	yes
	131.052859115354 	179.498809818896	M(C13)+H[1+]	-0.0004	-0.0	no
	152.031549466187 	179.529467326262	M+Na[1+]	-0.0003	-0.01	no
	198.038006551681 	173.861829437395	M+HCOONa[1+]	0.0007	-0.02	yes
E62 	Best guess: C00148, L-Proline	CE4788, C00763, C00148
	116.070559412872 	169.753632494793	M+H[1+]	-0.0	0.01	yes
	117.073760955606 	169.70894047301	M(C13)+H[1+]	-0.0002	0.01	yes
	138.05249818224 	169.360639746656	M+Na[1+]	-0.0001	0.01	no
E88 	Best guess: CE1554, N-acetyl-L-alanine	CE1554
	130.049715538965 	66.8188552456466	M[1+]	-0.0002	-0.0	no
	153.040390069727 	140.221429374424	M+Na[1+]	0.0007	0.02	yes
E304 	Best guess: C04299, D-myo-Inositol 1,2-cyclic phosphate	C04299
	243.0276125082 	210.87146136568	M+H[1+]	0.0011	0.02	yes
E240 	Best guess: C05829, 	C05829
	191.066363204313 	185.849912222947	M+H[1+]	0.0001	0.01	yes
E373 	Best guess: C04751, 5-amino-1-(5-phospho-D-ribosyl)imidazole-4-carboxylate	C04751
	340.054671978089 	169.736761012398	M+H[1+]	0.0006	0.03	yes
E336 	Best guess: C00294, Inosine	C00294
	269.088265932963 	146.088822290858	M+H[1+]	0.0002	-0.02	yes
	270.091189228949 	145.822689567745	M(C13)+H[1+]	-0.0003	-0.02	yes
	291.070333186397 	146.369284621043	M+Na[1+]	0.0003	-0.01	no
E170 	Best guess: C00956, L-2-Aminoadipate	C00956
	162.075890021504 	176.037706345584	M+H[1+]	-0.0002	0.02	yes
E305 	Best guess: C00140, N-Acetyl-D-glucosamine	C01074, C00645, C00140
	244.079324040822 	155.051494706048	M+Na[1+]	0.0001	0.01	yes
E210 	Best guess: C01586, Hippurate	CE5536, C01586
	180.065488342692 	36.7758046030458	M+H[1+]	0.0	-0.03	yes
E218 	Best guess: C00159, D-Mannose	C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	203.052613666493 	149.898814430292	M+Na[1+]	-0.0001	0.01	yes
E377 	Best guess: C00942, 3',5'-Cyclic GMP	C00942
	346.057788707732 	179.576020672889	M+H[1+]	0.0031	-0.01	yes
E91 	Best guess: C00408, L-Pipecolate	C05936, C00408
	130.086145983243 	156.843450542871	M+H[1+]	-0.0001	-0.03	yes
E339 	Best guess: C05401, 3-beta-D-Galactosyl-sn-glycerol	C05401
	277.089732447053 	173.184468027039	M+Na[1+]	0.0003	-0.01	yes
E127 	Best guess: C00819, D-Glutamine	C00064, C05100, C00819
	147.076381935189 	179.61347740865	M+H[1+]	0.0	-0.01	yes
	148.078935101464 	179.655294850676	M(C13)+H[1+]	-0.0008	-0.01	yes
	169.058297267761 	179.484874997502	M+Na[1+]	-0.0001	-0.01	yes
E92 	Best guess: C00408, L-Pipecolate	C05936, C00408
	130.086119695417 	188.119428816151	M+H[1+]	-0.0002	0.01	yes
	131.089245218966 	188.530128657552	M(C13)+H[1+]	-0.0004	0.0	no
E99 	Best guess: C00300, Creatine	C00300
	132.076578751122 	169.841525023698	M+H[1+]	-0.0002	0.04	yes
	133.079220685349 	169.832064371467	M(C13)+H[1+]	-0.001	0.05	yes
	154.05835318726 	170.397834572643	M+Na[1+]	-0.0004	0.03	yes
E128 	Best guess: C00819, D-Glutamine	C00064, C05100, C00819
	147.076484717754 	191.854864509991	M+H[1+]	0.0001	-0.01	yes
E101 	Best guess: C00407, L-Isoleucine	C00123, C00407
	132.101866089207 	153.538623115824	M+H[1+]	-0.0	0.02	yes
	133.104938127499 	153.424774861243	M(C13)+H[1+]	-0.0003	0.02	yes
E379 	Best guess: C00362, dGMP	C00020, C00362
	347.064966805791 	179.594738137385	M[1+]	0.0024	-0.01	yes
	348.067336232879 	179.591604585095	M+H[1+]	-0.003	-0.01	yes
E196 	Best guess: C00327, L-Citrulline	C00327
	175.095753701596 	187.732648435304	M[1+]	0.0006	-0.0	no
	176.102968551451 	182.294147749474	M+H[1+]	-0.0	-0.01	yes
	177.105626356365 	182.246445607546	M(C13)+H[1+]	-0.0008	-0.01	yes
	198.084927121057 	182.036220388696	M+Na[1+]	-0.0	-0.01	yes
E216 	Best guess: C00159, D-Mannose	C13747, C07480, C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	203.052650777453 	167.518683557958	M+Na[1+]	-0.0	0.01	yes
E29 	Best guess: C00133, D-Alanine	C00213, C00041, C00099, C00133
	90.055297651227 	170.125710486656	M+H[1+]	0.0003	0.02	yes
E214 	Best guess: C00159, D-Mannose	C13747, C07480, C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	181.071856704847 	49.9074854160341	M+H[1+]	0.0012	0.04	yes
	182.074876500163 	49.8573391766011	M(C13)+H[1+]	0.0008	0.04	yes
E215 	Best guess: C07480, 	C13747, C07480
	203.052605699518 	157.306679354031	M+Na[1+]	-0.0014	0.01	yes
	203.052613666493 	149.898814430292	M+Na[1+]	-0.0014	0.01	yes
	249.061228653683 	154.295838909063	M+HCOONa[1+]	0.0019	0.0	no
E102 	Best guess: C00152, L-Asparagine	C02642, C00152
	133.060695520522 	183.730017864071	M+H[1+]	-0.0001	-0.01	yes
	134.063340346426 	183.707904027794	M(C13)+H[1+]	-0.0008	-0.01	yes
	155.042554251059 	183.693592910717	M+Na[1+]	-0.0002	-0.01	yes
E70 	Best guess: C00183, L-Valine	C00719, C00431, C00183
	118.08619106547 	162.151489354211	M+H[1+]	-0.0001	0.01	yes
	119.089262402651 	164.636265573288	M(C13)+H[1+]	-0.0004	0.01	yes
	140.068068961457 	160.604592575579	M+Na[1+]	-0.0002	-0.01	yes
E37 	Best guess: C00082, L-Tyrosine	C00082, CE2174
	182.080751759631 	168.213651056991	M+H[1+]	-0.0004	0.02	yes
E217 	Best guess: C00159, D-Mannose	C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	181.07179618207 	37.2666769819766	M+H[1+]	0.0011	0.03	yes
E314 	Best guess: C00664, 5-Formiminotetrahydrofolate	C00664
	473.185636267121 	185.325552315865	M+H[1+]	-0.0035	0.04	yes
E282 	Best guess: C00864, (R)-Pantothenate	C00864
	220.118065127297 	59.5423480070644	M+H[1+]	0.0001	0.02	yes
E9 	Best guess: C05665, beta-Aminopropion aldehyde	C01888, C05665
	142.047326431032 	173.28763510354	M+HCOONa[1+]	-0.0002	-0.02	yes
	74.0608093047581 	173.098709903041	M+H[1+]	0.0007	-0.0	no
E105 	Best guess: C00402, D-Aspartate	C00049, C00402
	134.044614484966 	181.073887487692	M+H[1+]	-0.0002	0.02	yes
E211 	Best guess: C07480, 	CE5541, C13747, C07480
	180.065488342692 	36.7758046030458	M[1+]	0.0013	-0.03	yes
	181.07179618207 	37.2666769819766	M+H[1+]	-0.0002	0.03	yes
More data

Full data tables from pathway analysis, module analysis and activity network are stored in the 'tables/' directory. The tsv files can be imported into a spreadsheet program, e.g. MS Excel. Under the 'figures/network_modules/' directory are files intended for Cytoscape (cytoscape.org) visualization. Please refer to Cytoscape's guides for details. Details of this run are recorded in file mummichog.log.txt. 


# Results mummichog import

After running the algoritm (bash script on pc with py/mummichoc installed), 
todo into spectripy... !!!,
the following table os imported:

- userInput_to_EmpiricalCompounds.tsv

The first file is a copy of the data user gave to the algoritm, with the rowxxx identifiers needed as first column.
The second file gives the hypothetical annotations (kegg ID and comon name used in KEGG) for the retrieved metabolites.

please note that only metabolites which are present in KEGG can be retrieved using this approach.

#TODO!!! path, filename to auto select latest or in config?? or see git if name output cn change name/dir??

```{r load-mummichig-result-table}
#' load mummichog result files
DIR <- '1734448517.6233082.t2'    #umetab-bmi-lm
#DIR <- "1732118659.9126198.t2"   #test

list.dirs(path = DATA_INPUT_PATH, full.names = TRUE, recursive = TRUE)
list.dirs(path = DATA_OUTPUT_PATH, full.names = TRUE, recursive = TRUE)

DATA_OUTPUT_PATH <- paste0('/home/mdegraeve/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/', DIR, '/tables')
DATA_OUTPUT_PATH <- paste0(DATA_PATH, DIR, '/tables')


#' userInput_to_EmpiricalCompounds
handle <- file.path(DATA_OUTPUT_PATH, "userInput_to_EmpiricalCompounds.tsv")
userInput_to_EmpiricalCompounds <- read.csv(handle, header = TRUE, sep="\t")  |>
    as.data.frame()
userInput_to_EmpiricalCompounds[1:10, 1:10]

```

info regarding results file:

- adducts = str_row_ion
- keggid = compounds, also sometimes CE id from bigg)
- name = compound_names

*Remark 1*

one rowidxxx = compid = metabolite can have >1 matching kegg id, eg.

example:

#row957 E522 row957_M+H[1+];row970_M(C13)+H[1+] C00062;C00792
#L-Arginine$D-Arginine; D-2-Amino-5-guanidinovaleric acid

while in plot the label at E522 = L-Arginine

So.... want not these results as such but the table in the result.html report.
there only ones I want listed with 1 *best guess*

EmpiricalCompound	| Input m/z                     |	Retention time	| ion	           | mz_diff	| Statistic |	Significant
E522 	              Best guess: C00792, D-Arginine	                  C00062, C00792
	                  175.1183 	                      50.0	            M+H[1+]	        -0.0007	  -5.29	        yes
	                  176.1216 	                      39.0	            M(C13)+H[1+]  	-0.0008	  -23.32	      yes

2 options: C00062, C00792.
both are correct, one keggid is D- and other
L-form... so only know using bio background it is the L-form in mammals.

fyi: in activ netw edges w EC nrs
fyi: the '$' signs means ID annot is either x or y, eg L-Arginine$D-Arginine
	 the ';' is a synonym of x, eg D-2-Amino-5-guanidinovaleric acid
	 
solution:  keep all options into 1 rowidxxx


*Remark 2*

The entries with double rownames (1 compID with >1 match) are both possible too, but these are not
mentioned in the html report, and no idea how to know which one is correct.

example:

#row1020	E35	row1020_M+H[1+]	C13747;C07480;C00031;C00221;C02336;C00267;C00962;C00795;C00137;C00984;C00159	1,7-Dimethylxanthine; Paraxanthine$$D-Glucose$beta-D-Glucose$beta-D-Fructose; beta-Fruit sugar; beta-D-arabino-Hexulose; beta-Levulose; #Fructose$alpha-D-Glucose$beta-D-Galactose$D-Tagatose; lyxo-Hexulose$myo-Inositol$alpha-D-Galactose$D-Mannose	row1020	181.0712	502	0.165105022347	-1.45	AE_pos_181.0712_502
#row1020	E422	row1020_M+Na[1+]	CE4970;CE4968	2-methylbutyrylglycine$isovalerylglycine	row1020	181.0712	502	0.165105022347	-1.45	AE_pos_181.0712_502

so M+H for a hexose-weigth, OR M+Na for CExxx (not kegg id but CE4970te in BiGG models ID:
http://bigg.ucsd.edu/models/universal/metabolites/CE4970) =
2-methylbutyrylglycine$isovalerylglycine (isomers)

solution: want to keep both, merge into 1 rowidxxx



*Remark 3*

1 rowid, one emperical id, but lots of KEGG IDS (eg all hexoses), with analysis
best guess it is mannose. only in html report this decision is made.

example:

#row1239	E545	row1239_M+Na[1+]	C00031;C00221;C02336;C00267;C00962;C00795;C00137;C00984;C00159	D-Glucose$beta-D-Glucose$beta-D-Fructose; beta-Fruit sugar; beta-D-arabino-Hexulose; beta-Levulose; #Fructose$alpha-D-Glucose$beta-D-Galactose$D-Tagatose; lyxo-Hexulose$myo-Inositol$alpha-D-Galactose$D-Mannose	row1239	203.0517	59	2.67774722093E-06	-7.06	AE_pos_203.0517_59

E545 Best guess: C00159, D-Mannose 
C00031, C00221, C02336, C00267, C00962,
C00795, C00137, C00984, C00159 203.0517 59.0 M+Na[1+] -0.001 -7.06 yes 

based on module2 and activity network, there also upregulated n-acetl-glc,
n-acet-man (only indirectly...) and stachyose (gal-gal-glc-frc), myo-inositol
(glc-OH from g6p). so actually not so sure about this, best to leave all the
options as description.

alternatively, keep only the annotated ids with 1 hit (but losse the D-/l-aa and
-/l-sugars. so best to keep all info instead of picking best guess if i'm
doubting this was actually the best option.

based on m/z diff, Na-adduct and in mammals: row1239 = D-hexose.

TODO: maybe best manual go over reuslts.. (after mva p<0.05), go back to this
list and description change if needed

solution: keep all options into 1 rowidxxx (same adduct)


```{r cleanup-table-unique-rowids}
#' checks
#' total rowids (inlc duplic)
nrow(userInput_to_EmpiricalCompounds)
#' nr of non-duplic
length(userInput_to_EmpiricalCompounds$input_row[!duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))])
#' nr of duplic
length(unique(userInput_to_EmpiricalCompounds$input_row[duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))]))
#' needed length if correct merged == unique
length(unique(as.list(userInput_to_EmpiricalCompounds$input_row)))
#' nr of exactly 1 time occur in df
length(subset(userInput_to_EmpiricalCompounds$input_row, !(duplicated(userInput_to_EmpiricalCompounds$input_row) | duplicated(userInput_to_EmpiricalCompounds$input_row, fromLast = TRUE))))


#' names of the replicate rowids to be merged
doubles <- userInput_to_EmpiricalCompounds$input_row[duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))] #double hits, to merge into 1 rowid
doubles <- unique(doubles)

#' loop to make subdf with solved doubles
df_all <- NULL
i <- 1
for(i in 1:length(doubles)){
    #print(doubles[i])
        
    double_df <- userInput_to_EmpiricalCompounds[userInput_to_EmpiricalCompounds$input_row == doubles[i], ]

    #' add to empty df
    df <- NULL
    df$input_row <- double_df$input_row[1]
    df$EID <- paste(c(unlist(double_df$EID)),collapse = ';')
    df$str_row_ion <- paste(c(unlist(double_df$str_row_ion)),collapse = ';')
    df$compounds <- paste(c(unlist(double_df$compounds)),collapse = ';')
    df$compound_names  <- paste(c(unlist(double_df$compound_names)),collapse = ';')
    df$input_row.1 <- double_df$input_row[1]
    df$m.z  <- paste(c(unlist(double_df$m.z)),collapse = ';')
    df$retention_time  <- paste(c(unlist(double_df$retention_time)),collapse = ';')
    df$p_value  <- paste(c(unlist(double_df$p_value)),collapse = ';')
    df$statistic  <- paste(c(unlist(double_df$statistic)),collapse = ';')
    df$CompoundID_from_user <- paste(c(unlist(double_df$CompoundID_from_user)),collapse = ';')
    df <- as.data.frame(df)

    df_all <- rbind(df_all, df) 
}

#' checks
dim(df_all)
df_all[, 1:3]
colnames(df_all)
length(unique(as.list(df_all$input_row)))

#' subset original mummichog results table with rowid nr exactly 1 time occur in df
df_unique <- subset(userInput_to_EmpiricalCompounds, !(duplicated(userInput_to_EmpiricalCompounds$input_row) | duplicated(userInput_to_EmpiricalCompounds$input_row, fromLast = TRUE)))
dim(df_unique)
colnames(df_unique)

#' merge the non-duplic and the solved into 1 df again
clean_df <- rbind(df_unique, df_all)
clean_df <- clean_df[order(clean_df$input_row),]
dim(clean_df)
rownames(clean_df) <- clean_df$input_row

#' write for testing
handle <- file.path(DATA_PATH, 'cleanup_mummichog_output.txt')
write.table(clean_df, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

#fyi, top metab from lm FT01399 == glutamate according to mummichog

```

#todo!!!! see in clean df still some rowIDs merged in the untargeted metab df!!!


The computational 'EmpiricalCompound' and other information for a tentative
metabolite is now linked to the feature ID from the untargeted metabolomics
dataset. The candidate metabolite can group multiple ions, and be any of the
isobaric/isomeric species. See the original Mummichog 'results.html' for the
visualisation and for the EmpiricalCompounds prioritized by
pathway/module analysis.  

TODO: how add these to umetab dataset? using summExp object add description
maybe? FINDME!!! (todo after lcms semi quant untarg added, see there how).



# Acknowlegments




# Session information

```{r}
sessionInfo()

```
