# Creating the CHRIS untargeted LC-MS database(s)

In this document we define and create the *sqlite* database(s) with the full MS
data of the CHRIS untargeted metabolomics data. We create separate databases for
each polarity.

```{r}
library(readxl)
library(RSQLite)
library(MsBackendSql)
library(MsExperiment)
library(BiocParallel)
register(MulticoreParam(), 8L)

MZML_PATH <- "/var/run/media/jo/extdata/data/mzML"


```

Read the sample information.

```{r}
sd <- read_xlsx("data/chris_hilic_files.xlsx") |> as.data.frame()
```

## Positive polarity

Create a *MsBackendSQL* database with the full MS data.

```{r}
pos <- sd[sd$polarity == "POS", ]
fls <- file.path(MZML_PATH, pos$path, pos$file_name)
sqlite_db <- "data/chris_hilic_pos_1.0.0.3.sqlite"
con <- dbConnect(SQLite(), sqlite_db)
createMsBackendSqlDatabase(con, fls, MsBackendMzR(), chunksize = 16)
dbDisconnect(con)
```

Create a `MsExperiment` object *representing* the full CHRIS metabolomics
experiment.

```{r}
sps <- Spectra(sqlite_db, source = MsBackendOfflineSql(), drv = SQLite())
sps$file_name <- basename(dataOrigin(sps))
mse <- MsExperiment(spectra = sps, sampleData = pos)
mse <- linkSampleData(mse, with = "sampleData.file_name = spectra.file_name")
```

Store the sample information into the database.

```{r}
dbWriteSampleData(mse)
```

To load all data from the database.

```{r}
mse2 <- MsExperiment(
    spectra = Spectra(sqlite_db, source = MsBackendOfflineSql(),
                      drv = SQLite()))
```

## Negative polarity

Create a *MsBackendSQL* database with the full MS data.

```{r}
neg <- sd[sd$polarity == "NEG", ]
fls <- file.path(MZML_PATH, neg$path, neg$file_name)
sqlite_db <- "data/chris_hilic_neg_1.0.0.3.sqlite"
con <- dbConnect(SQLite(), sqlite_db)
createMsBackendSqlDatabase(con, fls, MsBackendMzR(), chunksize = 16)
dbDisconnect(con)
```

Create a `MsExperiment` object *representing* the full CHRIS metabolomics
experiment.

```{r}
sps <- Spectra(sqlite_db, source = MsBackendOfflineSql(), drv = SQLite())
sps$file_name <- basename(dataOrigin(sps))
mse <- MsExperiment(spectra = sps, sampleData = neg)
mse <- linkSampleData(mse, with = "sampleData.file_name = spectra.file_name")
```

Store the sample information into the database.

```{r}
dbWriteSampleData(mse)
```

To load all data from the database.

```{r}
mse2 <- MsExperiment(
    spectra = Spectra(sqlite_db, source = MsBackendOfflineSql(),
                      drv = SQLite()))
```