---
title: "General data overview and quality assessment of the CHRIS untargeted metabolomics data"
author: "Marilyn De Graeve, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-02-07"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

**Modified**: `r file.info("general_data_overview.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r biocstyle, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library(rmarkdown)
library(BiocStyle)
library(knitr)
library(kableExtra)
library(pander)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 10, fig.height = 7)
BiocStyle::markdown()
```

# Introduction

In this document, a general overview of the untargeted metabolomics data of the
Cooperative Health Research in South Tirol (CHRIS) study is provided. For a
description of the study, methods used for the collection, handling and
acquisition of the liquid chromatography-mass spectrometry (LC-MS) samples,
please see [@verri_hernandes_age_2022]. HILIC-based LC-MS data was measured for
about 8,000 serum samples from the CHRIS study over the time span of about 5
years. In each batch (which corresponds to one 96 well plate) data from 80 study
samples and a sample pool (QC POOL) was acquired. The same QC sample pool was
measured in each plate, always after 8 study samples. Samples are acquired in
both positive and negative ionization mode, the overview comprizes the positive
(pos) mode data only.


# Setup

## Directories

```{r settingsdefine paths, echo = FALSE}
#' General settings
filename <- "general_data_overview"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
DATA_PATH <- '/media/mdegraeve/josung/data/CHRIS' #'data' #FINDME!!!
```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(MsExperiment)
library(Spectra)
library(MsBackendSql)
library(RSQLite)
library(xcms)
library(pheatmap)
library(MsQuality)
library(readr)
library(readxl)
library(magick)
library(RColorBrewer)
suppressMessages(library(MetaboCoreUtils))
```

## Computation

```{r parallel_processing_setup, message=FALSE}
#' Set up parallel processing using multiple cores: get the number of
#' cpus allocated by queueing system or fall back to 7
NR_CORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 8)) - 1L
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(NR_CORES)) ###FINDME!!!
} else {
    register(bpstart(SnowParam(NR_CORES)))
}
```


# General import and exploration

## Data import

In the code block below we first load all required libraries and define the
location of the data.  The untargeted metabolomics data is stored within a
*SQLite* database that we load below as a `Spectra` object using the
`MsBackendOfflineSql` *backend* provided by the `r Biocpkg("MsBackendSql")`
package.

A sniplet of the Spectra object, the summary and columnames are shown below.


```{r load available data, echo = FALSE}
#' load other metabo_chris data, currently not used
#handle <- file.path(DATA_PATH, 'stds_biocrates.csv')
#stds_biocrates <- read_csv(handle, show_col_types = FALSE)
#head(stds_biocrates,5)
#colnames(stds_biocrates)

#handle <- file.path(DATA_PATH, 'chris_hilic_logbook.xlsx')
#chris_hilic_logbook <- read_excel(handle)
#head(chris_hilic_logbook,5)
#colnames(chris_hilic_logbook)


#' Get access to the SQLite database.
handle <- file.path(DATA_PATH, 'chris_hilic_pos_1.0.0.2.sqlite')
s <- Spectra(handle,
             source = MsBackendOfflineSql(),
             drv = SQLite())     #https://bioconductor.org/packages/release/bioc/vignettes/Spectra/inst/doc/Spectra.html

# info about structure S4 'Spectra' object
s
summary(s)
#str(s)

# colnames of spectra object
spectraVariables(s)

#stats per colnames
#summary(s$msLevel)
#summary(s$rtime)
#summary(s$polarity)
#max(s$scanIndex)
```

To summarize, one MS-level (`r unique(s$msLevel)`) is present, 1 polarity mode
(`r unique(s$polarity)`).
The retention time (rt) ranges from `r min(s$rtime)` - `r max(s$rtime)` seconds,
so an LC-MS method of approx `r round((max(s$rtime) - min(s$rtime))/60, 0)` min
was applied.  The scanIndex goes up to `r max(s$scanIndex)` scans.


The MS data (spectra) are next loaded as an `r Biocpkg("MsExperiment")` object,
which retrieves also the sample annotations for the data set from the SQL
database.  The `MsExperiment` object `chris` manages the *linkage* between
samples and spectra. The `length` of an `MsExperiment` is defined by the number
of samples (files) within the object.


```{r msExperiment object}
#' Load the data as an MsExperiment
#browseVignettes("MsExperiment")  #https://bioconductor.org/packages/release/bioc/manuals/MsExperiment/man/MsExperiment.pdf
chris <- MsExperiment(spectra = s)
chris

#nr of spectra
#' length(s)
```

So, there are `r length(chris)` samples and approx `r formatC(length(s), format
= "e", digits = 2)` MS1 spectra in the chris dataset. Still, the size of the
data set in memory is only `r print(object.size(chris), units = "MB")`


## General data overview

### Batches

In this section, an overview of the batches is given (number of samples (study,
  QC) per batch, date when batch was created etc.). For this, the function
  `sampleData(chris)` is used.


```{r sampleData(chris) batches, results = "asis"}
#' info about the sampledata
#head(sampleData(chris),5)
#ncol(sampleData(chris))
#nrow(sampleData(chris))
#colnames(sampleData(chris))

#' write to file for manuel inspection
#handle <- 'data/sampleData_chris.txt'
#write.table(sampleData(chris), file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

#' for each batch, count the bio and pool (and other sample types)
i <- 1
df3 <- data.frame()

for (i in 1:length(unique(sampleData(chris)$batch_id))) {
    #print(i)
    df2 <- NULL
    df2$batch_id <- unique(sampleData(chris)$batch_id)[i]
    df_batch <- sampleData(chris)[sampleData(chris)$batch_id == unique(sampleData(chris)$batch_id)[i], ]
    df2$date <- paste(df_batch$year[i], df_batch$month[i],
                      df_batch$day[i], sep='-')
    df2$nr_Pool <- nrow(df_batch[df_batch$sample_type == 'Pool', ])
    df2$nr_Study <- nrow(df_batch[df_batch$sample_type == 'Study', ])
    df2$plate_id <- df_batch$plate[1L]

    df3 <- rbind(df3, df2)
    }

rownames(df3) <- c(1:nrow(df3))
#print(df3)

kable_input <- kable(df3, caption = "Overview over batches and plates.")
kable_styling(kable_input, latex_options= c("scale_down", "hold_position"))
```

There are `r length(unique(sampleData(chris)$batch_id))` batches in the dataset.
The analysis of the metabo\_chris_untarget dataset took place over 5 years, with
data generated in 2016, 2017, 2020 and 2021.

During one batch, representing one day of analysis, there are approximately 80
biological samples (sample\_type == Study) and 10 pooled QCs run. The batch
starts and ends with internal QCs, in between each set of 8 consecutive study
samples, a QC was included. In one batch (BATCH0110 = BPLT00000207\_R), POOL\_1
-> POOL\_10 are named POOL\_12 -> POOL\_102, respectively. The order of the
samples remained the same however.  Also, the biological name in the
'file\_name' and 'sample\_id' didn't match for this batch.

This mismatch in BATCH0110 originates from a bug in the R code used to create
the sampleData: the trailing 2 for all "sample_id" came from the _R2 in the file
name. To solve in the SQLite database (performed only once), see code
bolow.

```{r fix bug SQLite DB trailing number, eval = FALSE}
#' run this once with R stated from the location of the DB
#' library(RSQLite)

#' connect to the database
#' con <- dbConnect(SQLite(), "chris_hilic_pos_1.0.0.2.sqlite")

#' get the sample annotations
#' sa <- dbGetQuery(con, "select * from sample_data")

#' identify samples from batch0110
#' idx <- which(sa$batch_id == "BATCH0110")

#' remove the trailing 2 from all sample_ids
#' sa$sample_id[idx] <- sub("2$", "", sa$sample_id[idx])

#' write the data back into the database
#' dbWriteTable(con, name = "sample_data", value = sa, overwrite = TRUE)
#' dbDisconnect(con)
```

Check to see is the samplenames of chris metadata are OK: yes.

```{r}
head(sampleData(chris)[sampleData(chris)$batch_id == 'BATCH0110',], 5)
```

For two plates (BPLT00000407 = BATCH0134 and BPLT00000411 = BATCH0155),
duplicated measurements are available. For the final data set, only one of the
two might be included. See Section [Study samples] for details.


### Study samples

In this section, information regarding the study samples is given (total number
of study samples, are there replicated study samples, etc.). Note: some plates
have been repeated because of failures or problems, so extra emphasis was given
to check whether this data is there.

Below we evaluate the total number of samples and unique samples.

```{r sampleData(chris) studysamples}
#' nr of batches
#' length(unique(sampleData(chris)$batch_id))

#' nr of unique biological samples
print('nr of unique biological samples:')
biolog_df <- sampleData(chris)[sampleData(chris)$sample_type == 'Study',]
length(unique(biolog_df$sample_id))

#' count non-unique overlapping biological samples
print('nr of overlapping biological samples:')
nrow(biolog_df) - length(unique(biolog_df$sample_id))
```

We next identify duplicated samples.

```{r}
#' duplicate biological samples
n_occur <- data.frame(table(biolog_df$sample_id))
#' n_occur[n_occur$Freq > 1,]

#' nice table of replicate bio samples
kable_input <- kable(n_occur[n_occur$Freq > 1,],
                     caption = "Study samples measured more than once.")
kable_styling(kable_input, latex_options= c("scale_down", "hold_position"))
```

No triplicates or other types of replicates (frequency > 2) etc samples are
detected.
Below we identify the batches and plates for these duplicated samples.

```{r}
#' filter sampleData(chris)
rep <- unlist(as.data.frame(n_occur[n_occur$Freq > 1,][1]) )
dupl_names <- as.vector(sapply(rep, as.character))
#' dupl_names[1]
replicates_df <- sampleData(chris)[sampleData(chris)$sample_id %in% dupl_names,]
#' nrow(replicates_df)
#' print(replicates_df)
print('batches with duplicate samples:')
unique(replicates_df$batch_id)

#' nice table of replicate bio samples
kable_input <- kable(
    replicates_df[, c("file_name", "plate", "time_stamp", "sample_id",
                      "batch_id", "injection_index")],
    caption = "Duplicated samples.")
kable_styling(kable_input, latex_options= c("scale_down", "hold_position"))

#' write to file for manuel inspection
#handle <- 'data/sampleData_chris_duplicates.txt'
#write.table(replicates_df, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)
```

The replicated samples represent indeed re-injected sample plates.

The duplicates are either present in both BATCH0155 (BPLT00000411) and
BATCH0185 (BPLT00000411\_R2); or in both BATCH0134 (BPLT00000407) and BATCH0135
(BPLT00000407_R). So, the latter batches are the same plate reanalyzed. See
filtered sampleData table "Duplicated samples" above.


## Data reorganisation

Batches BATCH0155 and BATCH0134 are removed, as the "chris\_hilic_logbook"
clearly states a valid reason.

- For plate BPLT00000411: "There was a clear lack of reproducibility over this
  batch. Needs repetition"
- For plate BPLT00000407: "First stage of sample preparation done in
  Feb/2020. Second stage and plate analysis performed in Nov/2020"

For some of the later batches, blank samples (solvent: unknown), QC Cal (pooled
QC samples in different concentrations), and QC ISDil (pooled QC with internal
standards spiked in different concentrations) samples have been added as well.

A summary of the number of samples for the various *sample types* are:

```{r, results = "asis"}
#' table with nr of study, pool and blank samples
table(sampleData(chris)$sample_type) |>
    as.data.frame() |>
    pandoc.table(style = "rmarkdown",
                 caption = "Numer of samples per sample type")
```

Only the study, QC and blank samples are retained at this point. The
available sample types are:

- *Blank*: blanks. TODO: need to figure out if it's pure water or
  solvent/matrix.
- *Pool*: pool of serum samples from about 5,000 participants of the CHRIS
  study.
- *Study*: serum samples from individual study participants.

```{r subset msexperiment}
#' Remove batch BATCH155 and BATCH134
chris <- chris[!sampleData(chris)$batch_id %in% c("BATCH0134", "BATCH0155")]
print('nr of samples without failed batches:')
nrow(sampleData(chris))

#' Restrict to Blank, Study and Pool samples
chris <- chris[sampleData(chris)$sample_type %in% c("Blank", "Pool", "Study")]
print('nr of samples without failed batches and without extra sample_types:')
nrow(sampleData(chris))
```

After subsetting the chris MSExperiment, `r nrow(sampleData(chris))` samples
remain in the study.

The number of samples for the various *sample types* are:

```{r, results = "asis"}
#' table with nr of study, pool and blank samples
table(sampleData(chris)$sample_type) |>
    as.data.frame() |>
    pandoc.table(style = "rmarkdown",
                 caption = "Numer of samples per sample type")
```


## Initial quality assessment

In this section, the initial quality of the samples is assessed.

- We view a sniplet of the sampleData and spectra objects.
- Samples are visualized using BPC and TIC, for a selection of samples of the
  data set, with different colors per batch.
- Plot the BPC and TIC separately for each batch. Whole dataset is used, with
  different colors per sample_type.
- Evaluation of chromatographic data between a selection of samples and the
  batches (binned TIC heat map).
- PCA of samples, colored by batch.
- The `MsQuality` Bioconductor package was checked.



Below a view of the sampleData of the chris MsExperiment with some random
samples.

```{r sampleData, echo=FALSE, results="asis"}
sampleData(chris)[, c(5, 10, 11, 13)] |>
    as.data.frame() |>
    head() |>
    pandoc.table(style = "rmarkdown",
                 caption = "Some samples from the data set.")
```

We also check the nr of spectra per sample.

```{r check nr of spectra per sample, echo=F}
#' get number of spectra per file     #mtcars |> sqrt() |> head() == head(sqrt(mtcars))
print('quantile nr spectra:')
df4 <- fromFile(chris) |>
    table()
quantile(df4)
#ok, nr of spetra per file are ok.

#doublecheck this, ok
#length(s) / min(df4)
#length(s) / nrow(sampleData(chris))  #s is spectra and chris is msexperiment
```

All samples in the experiment have (almost) the same amount of spectra: `r median(df4)` per
sample.  With `r nrow(sampleData(chris))` samples, this gives indeed a total of
`r length(s)` spectra in total.

To conclude, there are no problematic or failed samples (due to a low number of
spectra and thus a restricted retention time range.

Before the visualisation, filter the data to rt 10s - 250s using `filterRt`.
Before and after this rt window, no metabolites eluted and/or are measured.

```{r filter_rtime}
#' Filter for ret time
chris <- filterRt(chris, c(10, 250))

#' write to file for manuel inspection
#handle <- 'data/sampleData_chris_filtered_rtrange.txt'
#write.table(sampleData(chris), file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)
```

The retention time range for the entire data set is:
`r range(rtime(spectra(chris)))`


### BPC

For a few samples, the LC-MS data is visualized and a base peak chromatogram
(BPC) is plotted. For better view, the BPC is plotted colored by the different
batches.

```{r check qual with some samples and plot BPC}
#' Select POOL_4, POOL_5, POOL_6 from each batch.
chris_qc_sel <- chris[grep("^POOL_[4,5,6]", sampleData(chris)$sample_id)]
print('nr of samples in small subset of chris:')
length(chris_qc_sel)

#' Get the distribution of peak counts per file
#spectra(chris_qc_sel) |>
#    lengths() |>
#    split(fromFile(chris_qc_sel)) |>
#    lapply(quantile)

#' Calculate the distribution of total ion signal per file
#spectra(chris_qc_sel) |>
#    intensity() |>
#    sum() |>
#    quantile()

#quantile(sum(intensity(spectra(chris_qc_sel))))

#per file contains linear info per chrom of scannumber
#acquisitionNum(spectra(chris_qc_sel))
#scanIndex(spectra(chris_qc_sel))
#rtime(spectra(chris_qc_sel))


#' Define a unique color for each batch.
col_batch <- rainbow(length(unique(sampleData(chris)$batch_id)))
names(col_batch) <- unique(sampleData(chris)$batch_id)

#' Filter for ret time
chris_qc_sel <- filterRt(chris_qc_sel, c(10, 250))

#' Extract and plot a BPC
bpc <- chromatogram(chris_qc_sel, aggregationFun = "max")
plot(bpc,
     col = paste0(col_batch[bpc$batch_id], 20),
     main = "BPC after rt filtering")
grid()
legend("topright", inset=c(-0,0), col = col_batch, legend = names(col_batch),
       ncol = 3, lty = 1, cex=0.5)

#' save plot
png(paste0(IMAGE_PATH,'/chris_qc_sel_BPC.png'),
    width=11, height=7, units="in", res=150)
plot(bpc,
     col = paste0(col_batch[bpc$batch_id], 20),
     main = "BPC after rt filtering")
grid()
legend("topright", inset=c(-0,0), col = col_batch, legend = names(col_batch),
       ncol = 3, lty = 1, cex=0.5)
dev.off()
```

### TIC

For a few samples, the LC-MS data is visualized and a total ion chromatogram
(TIC) is plotted. For better view, the BPC is plotted colored by the different
batches.

```{r plot TIC}
#' Extract and plot a TIC
tic <- chromatogram(chris_qc_sel, aggregationFun = "sum")
plot(tic,
     col = paste0(col_batch[tic$batch_id], 20),
     main = "TIC after rt filtering")
grid()
legend("topright", inset=c(-0,0), col = col_batch, legend = names(col_batch),
       ncol = 3, lty = 1, cex=0.5)

#' save plot
png(paste0(IMAGE_PATH,'/chris_qc_sel_TIC.png'),
    width=11, height=7, units="in", res=150)
plot(tic,
     col = paste0(col_batch[tic$batch_id], 20),
     main = "TIC after rt filtering")
grid()
legend("topright", inset=c(-0,0), col = col_batch, legend = names(col_batch),
       ncol = 3, lty = 1, cex=0.5)
dev.off()
```

Clear differences between batches can be observed. Above BPC and TIC plots are
created with a selection of `r length(chris_qc_sel)` # samples.


### BPC and TIC per batch

Plot the BPC and TIC separately for each batch.  this is done on the whole chris
dataset.  In the notebook, the BPC and TIC are generated for all batches, see
the BPC and TIC folders for the images.  Code will be displayed for the first
one as an example but not for the others.

```{r plot BPC and TIC per batch, include=FALSE}
#' make directory for plots
dirBpc  <- file.path(IMAGE_PATH, 'BPC_per_batch')
dir.create(dirBpc, recursive = TRUE, showWarnings = FALSE)
dirTic  <- file.path(IMAGE_PATH, 'TIC_per_batch')
dir.create(dirTic, recursive = TRUE, showWarnings = FALSE)

#' color code according to sample_type
col_sampletype <- brewer.pal(8, "Accent")[c(1, 5, 8)]
names(col_sampletype) <- c("Study", "Pool", "Blank")


#' for each batch, generate BPC plot
i <- 1

for (i in 1:length(unique(sampleData(chris)$batch_id))) {
    #print(i)

    #' Select one batch.
    batch_id <- unique(sampleData(chris)$batch_id)[i]
    chris_batch <- chris[which(sampleData(chris)$batch_id == batch_id)]
    #length(chris_batch)

    #' match color to sample_type
    col_samples <- col_sampletype[sampleData(chris_batch)$sample_type]

    #' Extract and plot a BPC and TIC
    bpc <- chromatogram(chris_batch, aggregationFun = "max")
    tic <- chromatogram(chris_batch, aggregationFun = "sum")

    #' show plot BPC and TIC for 1st batch
    if (i == 1){
        plot(bpc,
             col = paste0(col_samples[bpc$sample_type], 60),
             main = paste0("BPC after rt filtering - ", batch_id))
        grid()
        legend("topright", inset=c(-0,0), col = col_sampletype,
               legend = names(col_sampletype),
               ncol = 1, lty = 1, cex=1)

        plot(tic,
             col = paste0(col_samples[tic$sample_type], 60),
             main = paste0("TIC after rt filtering - ", batch_id))
        grid()
        legend("topright", inset=c(-0,0), col = col_sampletype,
               legend = names(col_sampletype),
               ncol = 1, lty = 1, cex=1)

    }

    #' save plot for BPC and TIC
    png(paste0(dirBpc, '/chris_BPC_', batch_id, '.png'),
        width=11, height=7, units="in", res=150)
    plot(bpc,
         col = paste0(col_samples[bpc$sample_type], 60),
         main = paste0("BPC after rt filtering - ", batch_id))
    grid()
    legend("topright", inset=c(-0,0), col = col_sampletype,
           legend = names(col_sampletype),
           ncol = 1, lty = 1, cex=1)
    dev.off()

    png(paste0(dirTic, '/chris_TIC_', batch_id, '.png'),
        width=11, height=7, units="in", res=150)
    plot(tic,
         col = paste0(col_samples[tic$sample_type], 60),
         main = paste0("TIC after rt filtering - ", batch_id))
    grid()
    legend("topright", inset=c(-0,0), col = col_sampletype,
           legend = names(col_sampletype),
           ncol = 1, lty = 1, cex=1)
    dev.off()

}
```

Very dissimilar behaviour for all sample_types (POOL, Study and Blank samples)
in between the different batches are noted.

Some (subjective) observations for the BPC are listed below (TIC behave
similarly):

- Batch 0001 - 0004:
  - similar signal.
- Batch 0009 - 0024:
  - split of signal at around 30 seconds into multiple peaks.
  - decrease of intensity for peak around 140 seconds.
  - peaks after 150 seconds seem shifted to the right.
- Batch 0025 - 0028:
  - higher signal (noise/contamination?) between 50 and 125 seconds.
  - additional peak/signal at around 130 seconds.
  - generally higher signal.
- Batch 0036 - 0075:
  - higher signal/intensities for BP signal after 120 seconds.
  - see-saw signal between 75 and 125 seconds, that disappears after batch 0066.
- Batch 0079 - 0081:
  - general decrease in signal.
  - changes in peak composition at around 30 seconds.
- Batch 0087 - 0120:
  - peak composition at around 30 seconds like before.
  - low *noise* between 50 and 125 seconds.
- Batch 0125 - 0186:
  - loose signal in study samples from 20-30 seconds. Signal for QC samples
    still high.
  - high noise from 50 to 125 seconds.
  - decreasing signal intensity (with batch).


### Evaluation of signal for internal standards

To get an idea about the quality of the LC-MS chromatogram of the metabolic
features, the signals of the internal standards (IS) are evaluated.

The EICs for all the standards are generated by matching the *m/z* value followed
by extracting the signal for a relaxed rt window around the *expected* rt. The
lists with the *expected* rts has been curated manually thoroughly *a priori* by
Philippine Louail and Johannes Rainer; and are saved under the following
filenames:

- "Mix_alignment.txt"

This file contain the 29# standards which have been used for the alignment in
the NAFLD metabolomics experiment. 11# true IS (standards with heavy atoms 13C
and 15N) are present in this file, which were reported to have a good quality in
the positive ionisation mode (see file "internal_standard.txt" column
'quality_POS' for more info regarding the quality). The remaining 18# standards
are overlapping with the next file and will be extracted from the second file.

- "standards_dilution_nafld.txt"

This file contains the 60# (from a total of 200#) in-house standards which have
been acquired with a good quality in the positive ionisation mode. See column
'quality_POS' for more info regarding the quality.

We load and re-organize the files.

```{r load and cleanup standard lists}
##' list 1: Mix_alignment
#' load the IS list
mixAlignment_df <- read.delim(file.path(DATA_PATH, "Mix_alignment.txt"))
mixAlignment_df <- mixAlignment_df[!is.na(mixAlignment_df$POS), ]
rownames(mixAlignment_df) <- mixAlignment_df$abbreviation
#mixAlignment_df

#' keep only true IS
mixAlignment_df <- mixAlignment_df[grep('13C', mixAlignment_df$name),]
#nrow(mixAlignment_df)

#' clean up before merge (temp df)
df <- NULL
df$mix <- ''
df$name <- mixAlignment_df$name
df$abbreviation <- mixAlignment_df$abbreviation
df$HMDB.code <- ''
df$formula <- ''
df$POS <- mixAlignment_df$POS
df$NEG <- mixAlignment_df$NEG
df$RT <- mixAlignment_df$RT
df$data_set <- mixAlignment_df$data_set
df$sample <- mixAlignment_df$sample
df$operator <- mixAlignment_df$operator
df$version <- mixAlignment_df$version
df$quality_POS <- mixAlignment_df$quality_POS
df$quality_NEG <- mixAlignment_df$quality_NEG
df$note <- ''
df$mass <- mixAlignment_df$formula  #mass present here
df <- as.data.frame(df)
#df


##' list 2: standards_dilution_nafld
#' load the in-house standards list
standardsDilutionNafld_df <- read.delim(
    file.path(DATA_PATH, "standards_dilution_nafld.txt"))
standardsDilutionNafld_df <- standardsDilutionNafld_df[!is.na(standardsDilutionNafld_df$POS), ]
rownames(standardsDilutionNafld_df) <- standardsDilutionNafld_df$abbreviation
#standardsDilutionNafld_df
#nrow(standardsDilutionNafld_df)

#'calcualte neutral mass from formula
standardsDilutionNafld_df$mass <- mapply(standardsDilutionNafld_df$formula,
                                         FUN = calculateMass)   #https://bioconductor.org/packages/release/bioc/vignettes/MetaboCoreUtils/inst/doc/MetaboCoreUtils.html


##' merge
#' merge into standard list df
standards_chris_pos <- rbind(standardsDilutionNafld_df, df)
print('nr of (internal) standards included:')
nrow(standards_chris_pos)

#' calculate mz from neutral mass
standards_chris_pos$mass <- as.numeric(standards_chris_pos$mass)
standards_chris_pos$mz <- mapply(standards_chris_pos$mass,
                                 standards_chris_pos$POS,
                                 FUN = mass2mz)  #https://bioconductor.org/packages/release/bioc/vignettes/MetaboCoreUtils/inst/doc/MetaboCoreUtils.html


#' calc mz and rt region
standards_chris_pos$mzmin <- standards_chris_pos$mz - (standards_chris_pos$mz*10/1e6) #10ppm allow
standards_chris_pos$mzmax <- standards_chris_pos$mz + (standards_chris_pos$mz*10/1e6)
standards_chris_pos$rtmin <- standards_chris_pos$RT - 20
standards_chris_pos$rtmax <- standards_chris_pos$RT + 20


#' write to file for manuel inspection
#handle <- file.path(DATA_PATH, 'standards_chris_pos.txt')
#write.table(standards_chris_pos, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)
```

The extracted EICs for the IS are plotted per batch, as plotting all
`r nrow(sampleData(chris))` samples in one plot will be uninterpretable.

The EIC plot will be displayed one batch-standard combination as an example but
not for the others. See the EIC_IS folder for the images.

```{r plot EIC per batch, include=FALSE}
#' make directory for plots
dirEic  <- file.path(IMAGE_PATH, 'EIC_per_batch')
dir.create(dirEic, recursive = TRUE, showWarnings = FALSE)


#' color code according to sample_type
col_sampletype <- brewer.pal(8, "Accent")[c(1, 5, 8)]
names(col_sampletype) <- c("Study", "Pool", "Blank")


#' for each batch, generate EIC plots
i <- 1

for (i in 1:length(unique(sampleData(chris)$batch_id))) {
    #print(i)

    #' Select one batch.
    batch_id <- unique(sampleData(chris)$batch_id)[i]
    chris_batch <- chris[which(sampleData(chris)$batch_id == batch_id)]
    #length(chris_batch)

    dirBatch <- file.path(dirEic, batch_id)
    dir.create(dirBatch, showWarnings = FALSE)
    #' match color to sample_type
    col_samples <- col_sampletype[sampleData(chris_batch)$sample_type]

    #' for testing
    #chris_batch <- chris_batch[c(5, 10, 11, 13)] ###FINDME!!!
    #sampleData(chris_batch)

    #' Extract the EICs
    eics <- chromatogram(chris_batch,
                         mz = as.matrix(standards_chris_pos[, c("mzmin", "mzmax")]),
                         rt = as.matrix(standards_chris_pos[, c("rtmin", "rtmax")]),
                         chunkSize = NR_CORES)

    #iterate over standards
    for (j in seq_len(nrow(standards_chris_pos))) {
        #j <- 1

        standard_id <- standards_chris_pos$abbreviation[j]

        #' save plot for EIC
        png(paste0(dirBatch, '/chris_EIC_', batch_id, '_', standard_id, '.png'),
            width=11, height=7, units="in", res=150)
        plot(eics[j, ],
             col = paste0(col_samples[eics$sample_type], 80),
             main = paste0("EIC_", batch_id, '_', standard_id))
        grid()
        legend("topright", inset=c(-0,0), col = col_sampletype,
               legend = names(col_sampletype),
               ncol = 1, lty = 1, cex=1)
        mz_min_max <- fData(eics)[j, c("mzmin", "mzmax")]
        legend("topleft", legend = c(
                              standards_chris_pos$name[j],
                              paste0(c("m/z min: ", "m/z max: "),
                                     format(mz_min_max, digits = 8))))
        abline(v = standards_chris_pos$RT[j], col = "red", lty = 3)
        dev.off()

    }
}


#' show plot EIC for nth batch and nth standard as example
plot(eics[nrow(standards_chris_pos),],
     col = paste0(col_samples[eics$sample_type], 80),
     main = paste0("EIC_", batch_id, '_', standard_id))
grid()
legend("topright", inset=c(-0,0), col = col_sampletype,
       legend = names(col_sampletype),
       ncol = 1, lty = 1, cex=1)
abline(v = standards_chris_pos$RT[j], col = "red", lty = 3)

```


### Evaluation of chromatographic data between samples and batches

We next aim to explore similarities and dissimilarities between the general
chromatographic data between samples and batches. We thus compare the total ion
signal between the various measurements for a few samples of each batch, binned
by 2 seconds.


```{r heatmap intra corr samples tic, fig.height=8, fig.width=7}
#' Select POOL_4, POOL_5, POOL_6 from each batch.
chris_qc_sel <- chris[grep("^POOL_[1,2,3,4,5,6,7,8,9,10]", sampleData(chris)$sample_id)]  #!!!FINDME
print('nr of samples in small subset of chris:')
length(chris_qc_sel)

#' Heatmap from total ion chromatogram
tic <- chromatogram(chris_qc_sel, aggregationFun = "sum", chunkSize = NR_CORES) |>
    bin(binSize = 2)

#' Correlation between the binned TICs.
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
    cor()

#' yes the code is weird but pheatmap has a weird  problems
#' with rownames/colnames handling.
col_hm <- data.frame(sample_type = sampleData(chris_qc_sel)[, "sample_type"])
rownames(col_hm) <- colnames(ticmap)
row_hm <- data.frame(batch = sampleData(chris_qc_sel)[, "batch_id"])
rownames(row_hm) <- rownames(ticmap)
ann_color <- list(
    sample_type = col_sampletype,
    batch = col_batch
)

rownames(ticmap) <- rownames(row_hm)
colnames(ticmap) <- rownames(col_hm)

#' plot heatmap
pheatmap(ticmap, annotation_row = row_hm, annotation_col = col_hm,
         annotation_colors = ann_color, annotation_names_row = FALSE,
         annotation_names_col = FALSE, show_rownames = FALSE,
         show_colnames = FALSE, annotation_legend = TRUE)

```

With full legend batches for better readability:

```{r }
par(mar=rep(0, 4), xpd = NA)
plot(rbind(c(1,745),c(962,1)),bty ="n",axes=F,frame.plot=F, xaxt='n', ann=FALSE, yaxt='n', asp=745/962, cex = 0.1, col = 'white')
legend("topright", inset=c(-0,0), col = col_batch, legend = names(col_batch),
       ncol = 3, lty = 1, cex=.8)

```

As expected, QC samples group mostly by batch. The blank
samples and biological samples are not present to be evaluated.

### PCA of samples

Unsupervised PCA is shown below on subset of samples, to have an idea of sample
clustering, colored per batch.

```{r pca samples, fig.height=8, fig.width=7}
#' for testing
#set.seed(1)
#chris_sel <- chris[sample(1:length(chris), size = 10, replace = TRUE)] ###FINDME!!!
#sampleData(chris_sel)
chris_qc_sel <- chris[grep("^POOL_[4,5,6]", sampleData(chris)$sample_id)]
print('nr of samples in small subset of chris:')
length(chris_qc_sel)

#' calc TIC per sample
tic <- chromatogram(chris_qc_sel, aggregationFun = "sum", chunkSize = NR_CORES) |>
    bin(binSize = 2)

#' create matrix with intensities (samples*binned TIC variables).
ticmap <- do.call(cbind, lapply(tic, intensity))

#' Define a unique color for each batch.
col_batch <- rainbow(length(unique(sampleData(chris)$batch_id)))
names(col_batch) <- unique(sampleData(chris)$batch_id)

#' PCA of tic samples, colored by batch
pca_samples <- prcomp(ticmap)
scores = as.data.frame(pca_samples$x)
plot(x = scores$PC1, y = scores$PC2,
     col = paste0(col_batch[tic$batch_id], 80),
     pch = 19,
     main = "PCA scoreplot with binned TIC")
grid()
#legend("topright", inset=c(-0,0), col = col_batch, legend = names(col_batch),
#       ncol = 3, lty = 1, cex=0.05)

#' save plot
png(paste0(IMAGE_PATH,'/chris_qc_sel_PCA.png'),
    width=11, height=7, units="in", res=150)
plot(x = scores$PC1, y = scores$PC2,
     col = paste0(col_batch[tic$batch_id], 80),
     pch = 19,
     main = "PCA scoreplot with binned TIC")
grid()
dev.off()

```



### MsQuality

The `MsQuality` Bioconductor package was used to evaluate some of the
pre-defined quality metrics.

```{r msquality }

#' check msquality with spectra object
#' https://bioconductor.org/packages/release/bioc/vignettes/MsQuality/inst/doc/MsQuality.html

#' best subset s first, how?? TODO

#' see options
print('MsQuality metric options:')
qualityMetrics(s)

#calculateMetrics(s, metrics = qualityMetrics(s), filterEmptySpectra = TRUE,
#    relativeTo = "Q1", msLevel = 1L)
```


```{r save Rdata}
save(chris, file = file.path(RDATA_PATH, "chris_after_visu.RData"))
```


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```

# References
