---
title: "General data overview and quality assessment of the CHRIS untargeted metabolomics data"
author: "Marilyn De Graeve, Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
---

**Modified**: `r file.info("general_data_overview.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r biocstyle, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library(BiocStyle)
library(knitr)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 10, fig.height = 7)
BiocStyle::markdown()
```

```{r settings, echo = FALSE}
#' General settings
filename <- " general_data_overview"
#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)
```

# Introduction

In this document we provide a general overview of the CHRIS untargeted
metabolomics data. HILIC-based LC-MS data was measured for about 7,000 serum
samples from the CHRIS study over the time span of about 5 years. In each batch
(which corresponds to one 96 well plate) data from 80 study samples and a sample
pool (QC POOL) was acquired. The same QC sample pool was measured in each plate,
always after 8 study samples.

# General data overview and initial quality assessment

The untargeted metabolomics data is stored within a *SQLite* database that we
load below using the `r Biocpkg("MsBackendSql")` package.

```{r}
#' Load libraries
library(MsExperiment)
library(Spectra)
library(MsBackendSql)
library(RSQLite)
library(xcms)
library(pheatmap)
library(MsQuality)

#' Get access to the SQLite database.
s <- Spectra("data/chris_hilic_pos_1.0.0.2.sqlite",
             source = MsBackendOfflineSql(),
             drv = SQLite())
#' Load the data as an MsExperiment
chris <- MsExperiment(spectra = s)
```


## General overview

- Overview of batches: number of samples (study, QC) per batch. Date when batch
  was created etc. use `sampleData(chris)` for that.
- Total number of study samples, are there replicated study samples? note: some
  plates were repeated because of failures or problems, maybe also check if the
  data is there.

## Initial quality assessment

- maybe have a look at the `MsQuality` Bioconductor package to use one of the
  pre-defined quality metrics?
- check problematic or failed samples (low number of spectra? restricted
  retention time range?).
- maybe BPC or TIC (of only QC samples) of the data set, different colors per
  batch.
- heatmap comparing BPCs.
- boxplots with distribution of


# Session information

```{r}
sessionInfo()
```