---
title: "XCMS processing of the untargeted metabolomics data from CHRIS"
author: "Mar Garcia-Aloy, Johannes Rainer"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
---

```{r biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown()
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

**Modified**: `r file.info("peak_detection.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = TRUE, results = "hide", message = FALSE}
## Set general options
options(useFancyQuotes = FALSE)
set.seed(123)

## Define paths:
filename <- "peak_detection"
## Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE, showWarnings = FALSE)
## Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

exclusion <- c("041016_BLANK_POS_1",
               "041016_BLANK_POS_2-041016_BLANK_POS_1",
               "041016_BLANK_POS_2",
               "041016_BLANK_POS_1 (3)",
               "041016_BLANK_POS_1 (4)",
               "06102032_BLANK_POS_b",
               "251016_BLANK_POS_1",
               "251016_BLANK_POS_2",
               "20161025_BLANK_POS_a",
               "20161025_BLANK_POS_b",
               "211116_BLANK_POS_1",
               "211116_BLANK_POS_2",
               "051216_BLANK_POS_1",
               "051216_BLANK_POS_2",
               "20161205_BLANK_POS_a",
               "20161205_BLANK_POS_b",
               "20161205_BLANK_POS_a_R",
               "20161205_BLANK_POS_a_R (2)",
               "20161205_BLANK_POS_b_R",
               "20161205_BLANK_POS_a_R1",
               "090117_BLANK_POS_1",
               "090117_BLANK_POS_2",
               "20170112_BLANK_POS_a",
               "20170112_BLANK_POS_b",
               "120117_BLANK_POS_1",
               "20170209_BLANK_POS_a",
               "20170406_BLANK_POS_a",
               "20170731_BLANK_POS (2)",

               "06102016_0010060146_POS",
               "20170615_0010282264_POS_R",

               "20161116_POOL_POS_2",
               "20170405_0010007645_POS",

               "20170209_BLANK_NEG_a",
               "20170209_BLANK_NEG_b",
               "20170220_BLANK_NEG_a",
               "20170220_BLANK_NEG_a (2)",
               "20170224_BLANK_NEG_a",
               "20170407_BLANK_NEG_a",
               "20170731_BLANK_NEG",
               "20170109_BLANK_NEG_a",
               "20170111_BLANK_NEG_a",
               "20170206_BLANK_NEG_a",
               "20170207_BLANK_NEG_a",

               "20161024_0010302963_NEG",
               "20161201_POOL_NEG_1.mzML",
               "20161201_POOL_NEG_2.mzML")

## Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 6))

rt_cut <- 340

MZML_PATH <- "/data/massspec/mzML/"
if (!dir.exists(MZML_PATH))
    stop("Can not find the directory with the mzML files: ", MZML_PATH)

```

# Introduction

In this document we perform the peak detection of the untargeted metabolomics
data of the CHRIS population study, both for positive and negative polarity.

We first load and define the phenodata table with all samples/mzML files which
are part of the present data set.

```{r libraries, message = FALSE}
library(readxl)
library(RColorBrewer)
library(xcms)

injections <- as.data.frame(read_xlsx("data/chris-files-annotated.xlsx"))
injections$id <- gsub(".mzML", "", injections$mzML_file)
injections <- injections[order(injections$timestamp), ]
injections$batch <- injections$run_id
injections <- injections[!injections$sample_name %in% exclusion, ]
injections <- injections[!injections$type %in% c("ACN_IS", "EQ_MIX"), ]

## Define the injection index
runids <- split(injections$run_id,
                f = factor(injections$run_id,
                           levels = unique(injections$run_id)))
injections$inj_idx <- unlist(lapply(lengths(runids), seq), use.names = FALSE)

## split by polarity
pd_pos <- injections[injections$polarity == "POS", ]
pd_neg <- injections[injections$polarity == "NEG", ]
```

Performing first the analysis for the positive polarity data. Negative polarity
data will be processed afterwards using the same settings. Below we load the
positive polarity data, subset to the defined retention time range and remove
unnecessary spectrum metadata.

```{r pos-load}
data_pos <- readMSData(files = paste0(MZML_PATH, pd_pos$mzML_file),
                       pdata = as(AnnotatedDataFrame(pd_pos),
                                  "NAnnotatedDataFrame"),
                       mode = "onDisk")

data_pos <- filterRt(data_pos, c(0, rt_cut))
data_pos <- selectFeatureData(
    data_pos, fcol = c("fileIdx", "spIdx", "acquisitionNum", "retentionTime",
                       "msLevel", "precursorScanNum", "polarity", "centroided",
                       "seqNum")
)
```

# Peak detection and refinement

We perform a centWave-based peak detection followed by a *peak refinement* step
that aims at reducing the peak detection artifacts, such as split peaks or
overlapping peaks. Below we define the parameters for the peak detection and
subsequently perform the analysis.

```{r pos-peak-detection, warning = FALSE, message = FALSE}
cwp <- CentWaveParam(
    peakwidth = c(2, 20),
    ppm = 50,
    snthresh = 5,
    mzdiff = 0.001,
    prefilter = c(3, 500),
    noise = 100,
    integrate = 2)

register(MulticoreParam(ncores))
tm <- system.time(
    data_pos <- findChromPeaks(data_pos, param = cwp)
)
save(data_pos, file = paste0(RDATA_PATH, "data_pos.RData"))
```

Next we refine the identified chromatographic peaks. Processing took
`r tm["elapsed"] * 60` minutes.

```{r pos-peak-refinement, warning = FALSE, message = FALSE}
mnp <- MergeNeighboringPeaksParam(
    expandRt = 2,
    expandMz = 0.001,
    ppm = 10,
    minProp = 0.66)

## register(MulticoreParam(workers = ceiling(ncores * 2/3),
##                         timeout = 86400L))
tm <- system.time(
    data_pos <- refineChromPeaks(data_pos, param = mnp, BPPARAM = bpparam())
)
save(data_pos, file = paste0(RDATA_PATH, "data_pos_ref.RData"))
```

Processing took `r tm["elapsed"] * 60` minutes. At last we evaluate the number
of detected peaks per sample and plate as well as sample type.

```{r pos-peaks-per-file, fig.path = IMAGE_PATH, caption = "Positive polarity data: peak count and signal distribution per sample.", fig.width = 16, fig.height = 10, echo = FALSE}

col_type <- brewer.pal(7, "Set1")[c(2, 3, 5)] # blue, green, orange
names(col_type) <- c("study", "QC", "BLANK")

intos <- split(log2(chromPeaks(data_pos)[, "into"]),
               f = as.integer(chromPeaks(data_pos)[, "sample"]))
intos_mean <- lapply(intos, mean)

## pdf("peak_summary.pdf", width = 12, height = 8)

par(mfrow = c(2, 1), mar = c(0.5, 4.5, 0.5, 0.5))
barplot(lengths(intos), border = paste0(col_type[data_pos$type], 40),
        ylab = "peak count", xaxt = "n", main = "positive polarity")
boxplot(intos, outline = FALSE, medlty = 0, whisklty = 0, staplelty = 0,
        border = paste0(col_type[data_pos$type], 40), ylim = c(10, 15),
        xaxt = "n", ylab = expression(log[2]~intensity))
points(x = seq_along(intos_mean), intos_mean,
       col = paste0(col_type[data_pos$type], 80))

year_month <- paste(data_pos$year, data_pos$month, sep = "-")
is <- table(year_month)
year_month_x <- c(1, cumsum(is))
for (i in 2:length(year_month_x)) {
    arrows(x0 = year_month_x[i - 1], x1 = year_month_x[i],
           y0 = 14.5, y1 = 14.5, code = 3, angle = 90, length = 0.1)
    text(x = mean(year_month_x[c(i - 1, i)]), y = 14.9,
         label = names(year_month_x)[i])
}

## dev.off()

```

While the number of detected peaks is comparable within batch (plate) a large
differences can be observed between some batches. A comparably low number of
peaks was identified in the first 4 batches, while the largest number of
detected peaks are present in the last batches. In contrast to the number of
peaks, the intensity of the detected peaks is comparable across all batches.

```{r neg-polarity}
data_neg <- readMSData(files = paste0(MZML_PATH, pd_neg$mzML_file),
                       pdata = as(AnnotatedDataFrame(pd_neg),
                                  "NAnnotatedDataFrame"),
                       mode = "onDisk")

data_neg <- filterRt(data_neg, c(0, rt_cut))
data_neg <- selectFeatureData(
    data_neg, fcol = c("fileIdx", "spIdx", "acquisitionNum", "retentionTime",
                       "msLevel", "precursorScanNum", "polarity", "centroided",
                       "seqNum")
)
## register(MulticoreParam(ncores))
system.time(
    data_neg <- findChromPeaks(data_neg, param = cwp)
)
save(data_neg, file = paste0(RDATA_PATH, "data_neg.RData"))
## Peak refinement
## register(MulticoreParam(workers = ceiling(ncores * 2/3),
##                         timeout = 86400L))
system.time(
    data_neg <- refineChromPeaks(data_neg, param = mnp, BPPARAM = bpparam())
)
save(data_neg, file = paste0(RDATA_PATH, "data_neg_ref.RData"))
```

```{r neg-peaks-per-file, fig.path = IMAGE_PATH, caption = "Negative polarity data: peak count and signal distribution per sample.", fig.width = 16, fig.height = 10, echo = FALSE}

col_type <- brewer.pal(7, "Set1")[c(2, 3, 5)] # blue, green, orange
names(col_type) <- c("study", "QC", "BLANK")

intos <- split(log2(chromPeaks(data_neg)[, "into"]),
               f = as.integer(chromPeaks(data_neg)[, "sample"]))
intos_mean <- lapply(intos, mean)

## pdf("peak_summary.pdf", width = 12, height = 8)

par(mfrow = c(2, 1), mar = c(0.5, 4.5, 0.5, 0.5))
barplot(lengths(intos), border = paste0(col_type[data_neg$type], 40),
        ylab = "peak count", xaxt = "n", main = "negative polarity")
boxplot(intos, outline = FALSE, medlty = 0, whisklty = 0, staplelty = 0,
        border = paste0(col_type[data_neg$type], 40), ylim = c(10, 15),
        xaxt = "n", ylab = expression(log[2]~intensity))
points(x = seq_along(intos_mean), intos_mean,
       col = paste0(col_type[data_neg$type], 80))

year_month <- paste(data_neg$year, data_neg$month, sep = "-")
is <- table(year_month)
year_month_x <- c(1, cumsum(is))
for (i in 2:length(year_month_x)) {
    arrows(x0 = year_month_x[i - 1], x1 = year_month_x[i],
           y0 = 14.5, y1 = 14.5, code = 3, angle = 90, length = 0.1)
    text(x = mean(year_month_x[c(i - 1, i)]), y = 14.9,
         label = names(year_month_x)[i])
}

## dev.off()

```


# Session information

```{r}
sessionInfo()
```
